# A worked record-linkage pipeline {#sec-workflow}

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

This chapter strings every piece from @sec-quickstart together into a
single, realistic linkage pipeline: an incoming batch of records needs to
be matched against a reference database. It's the same shape of problem
`examples_data_linkage.R` in the repo demonstrates, written out here as a
narrative.

## The data

```{r}
#| label: data
incoming <- data.frame(
    id   = 1:4,
    name = c("  JOHN O'BRIEN  ", "MARY-JANE SMITH", "ROBERT BROWN", "Kyle John Haynes"),
    dob  = as.Date(c("1985-03-15", "1990-07-22", "1975-12-01", "1988-01-05")),
    stringsAsFactors = FALSE
)

reference <- data.frame(
    ref_id = 101:105,
    name   = c("JON OBRIEN", "MARYANN SMITH", "ROB BROWN", "JOHN BRIEN", "Haynes John Kyle"),
    dob    = as.Date(c("1985-03-14", "1990-07-20", "1975-12-02", "1985-03-15", "1988-01-05")),
    stringsAsFactors = FALSE
)

incoming
reference
```

Four incoming records, five reference records, deliberately messy:
inconsistent casing, stray whitespace, punctuation, typos, one straight-up
near-duplicate (`"JOHN BRIEN"` vs. `"JOHN O'BRIEN"`), and one fully
reordered multi-token name (`"Kyle John Haynes"` vs.
`"Haynes John Kyle"`).

## Step 1: normalise

Strip punctuation and whitespace inconsistencies *before* comparing, so the
comparison step measures genuine name similarity rather than formatting
noise.

```{r}
#| label: normalise
incoming$name_clean  <- fchartr("'-", "  ", ftrimws(incoming$name))
reference$name_clean <- fchartr("'-", "  ", ftrimws(reference$name))

incoming$name_clean
reference$name_clean
```

(`jaro_winkler_tokens()` does its own internal punctuation stripping and
case-folding — see @sec-reference — but normalising up front keeps this
example explicit, and matters for `jaro_winkler()`/`jaro_winkler_matrix()`,
which don't.)

## Step 2: block

Comparing every incoming record against every reference record is
$O(n \times m)$ — fine at this toy scale, prohibitive at millions of rows.
**Blocking** groups records by a cheap, coarse key first (here: a
month-year DOB key) and only compares records *within* the same block,
cutting the number of pairwise comparisons by orders of magnitude in
practice.

```{r}
#| label: blocking
incoming$dob_block  <- fsubstr(format_date(incoming$dob,  "compact"), 1, 6)
reference$dob_block <- fsubstr(format_date(reference$dob, "compact"), 1, 6)

incoming[, c("id", "name", "dob_block")]
reference[, c("ref_id", "name", "dob_block")]
```

Soundex/NYSIIS codes (@sec-theory) are the other common blocking key,
typically used *together* with a coarse date block — block on
`soundex(last_name)` + birth year, for instance — to keep block sizes
manageable while still catching records where the date itself has a typo.

## Step 3: score within each block

For each incoming record, find reference records sharing its block, then
score name similarity only among those candidates. `jaro_winkler_tokens()`
is used uniformly rather than switching between it and plain
`jaro_winkler()`: it degenerates exactly to ordinary Jaro-Winkler when both
names happen to have the same token count and order (@sec-theory), so
there's no accuracy cost on the "normal" records — and it correctly
survives the one record (`"Kyle John Haynes"`) where the tokens are fully
reordered, which plain `jaro_winkler()` would score badly.

```{r}
#| label: scoring
match_one <- function(rec) {
    candidates <- reference[reference$dob_block == rec$dob_block, ]
    if (nrow(candidates) == 0L) return(NULL)
    scores <- jaro_winkler_tokens(rep(rec$name_clean, nrow(candidates)), candidates$name_clean)
    candidates$score <- scores
    candidates[order(-candidates$score), c("ref_id", "name", "score")]
}

results <- lapply(seq_len(nrow(incoming)), function(i) {
    res <- match_one(incoming[i, ])
    if (is.null(res)) return(NULL)
    cbind(incoming_id = incoming$id[i], incoming_name = incoming$name[i], res)
})
do.call(rbind, results)
```

## Step 4: threshold

Apply a similarity cutoff to decide which candidate pairs are accepted as
links. 0.85 is a common starting point for Jaro-Winkler on names; the
right value depends on your data and how costly false positives/negatives
are relative to each other — tune it against a labelled sample if you
have one.

```{r}
#| label: threshold
all_results <- do.call(rbind, results)
threshold <- 0.85
links <- all_results[all_results$score >= threshold, ]
links[!duplicated(links$incoming_id), ]  # best link per incoming record
```

Every incoming record links to its obvious match, including the fully
reordered `"Kyle John Haynes"` → `"Haynes John Kyle"` pair — the one case
in this dataset where plain `jaro_winkler()` would have scored well below
threshold and `jaro_winkler_tokens()` is doing real work, not just
matching what character-level comparison already would have.

@sec-benchmarks now scales the comparison step of this exact pipeline
shape up to 3 million rows.
