# Function reference {#sec-reference}

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

Every exported function, grouped the same way as the README. Each entry
gives the signature, every argument, the return value, and at least one
runnable example. This chapter is the detailed reference; @sec-quickstart
is the faster tour.

## Matching & substitution

### `fgrepl()`

Equivalent to `base::grepl()`, using PCRE2 and Intel TBB. Large inputs can
use multiple cores; the crossover depends on subject length and pattern
cost. See @sec-theory for the automatic PCRE-only-syntax fallback to
`base::grepl(perl = TRUE)`.
With `fixed = TRUE`, the byte-oriented pattern is prepared once: one-byte
needles use `memchr`, two- and three-byte needles use `memchr` plus
`memcmp`, and longer needles use Boyer–Moore–Horspool. `ignore.case = TRUE`
uses a 256-entry byte-folding table without lowercasing each input string.

```r
fgrepl(pattern, x, ignore.case = FALSE, perl = FALSE,
       fixed = FALSE, useBytes = FALSE, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `pattern` | Character scalar. Pattern to search for. |
| `x` | Character vector. `NA` elements return `NA`. |
| `ignore.case` | Logical. Case-insensitive matching. |
| `perl` | Logical. If `TRUE`, skip the PCRE-only syntax check and run the parallel PCRE2 engine directly. |
| `fixed` | Logical. Treat `pattern` as bytes and use the prepared literal-search path. |
| `useBytes` | Logical. Ignored; kept for signature compatibility with `base::grepl()`. |
| `nthreads` | Positive integer per-call thread cap, or `NULL` for the RcppParallel default; `1` forces serial execution. |

**Returns:** logical vector the same length as `x`.

```{r}
fgrepl("^inv", c("invoice", "Invoice", "credit"), ignore.case = TRUE)
fgrepl("a.b", c("axb", "ab"), fixed = TRUE)  # literal "a.b" — no match for "axb"
```

### `fgrep()`

Equivalent to `base::grep()`: wraps `fgrepl()` and returns indices or
values.

```r
fgrep(pattern, x, ignore.case = FALSE, perl = FALSE, value = FALSE,
      fixed = FALSE, useBytes = FALSE, invert = FALSE, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `pattern`, `x`, `ignore.case`, `perl`, `fixed`, `useBytes`, `nthreads` | Same as `fgrepl()`. |
| `value` | Logical. Return matching elements instead of indices. |
| `invert` | Logical. Return non-matching indices/values instead. |

**Returns:** integer vector of indices, or a character vector when
`value = TRUE`.

```{r}
fgrep("^inv", c("invoice", "credit", "invoice2"))
fgrep("^inv", c("invoice", "credit", "invoice2"), value = TRUE)
```

### `fsub()`

Equivalent to `base::sub()`: first-match substitution. Supports `\\1`-`\\9`
capture groups and `\\U`/`\\L`/`\\E` case conversion in `replacement`.

```r
fsub(pattern, replacement, x, ignore.case = FALSE, perl = FALSE,
     fixed = FALSE, useBytes = FALSE, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `pattern` | Character scalar. Pattern to search for. |
| `replacement` | Character scalar. Replacement string. |
| `x` | Character vector. `NA` elements return `NA`. |
| `ignore.case`, `perl`, `fixed`, `useBytes`, `nthreads` | Same as `fgrepl()`. |

**Returns:** character vector the same length as `x`.

```{r}
fsub("-([0-9]+)$", "_\\1", c("invoice-001", "invoice-002"))
```

### `fgsub()`

Equivalent to `base::gsub()`: global substitution. Same capture-group and
case-conversion support as `fsub()`.

```r
fgsub(pattern, replacement, x, ignore.case = FALSE, perl = FALSE,
      fixed = FALSE, useBytes = FALSE, nthreads = NULL)
```

Arguments are identical to `fsub()`. **Returns:** character vector the
same length as `x`.

```{r}
fgsub("[aeiou]", "_", c("hello world", "fast string"), ignore.case = TRUE)
```

### `gsub_all()`

No base R equivalent: applies several patterns in a single function call,
either sequentially (chained, like repeated `fgsub()` calls — later
patterns can match text introduced by earlier replacements) or as one
combined left-to-right scan (`sequential = FALSE` — the first pattern to
match at each position wins; later replacements never see earlier ones'
output).

```r
gsub_all(patterns, replacements, x, fixed = FALSE, ignore.case = FALSE,
          sequential = TRUE, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `patterns` | Character vector of patterns to search for. |
| `replacements` | Character scalar or vector the same length as `patterns`. |
| `x` | Character vector. `NA` elements return `NA`. |
| `fixed` | Logical. Treat each pattern as a literal string. |
| `ignore.case` | Logical. Case-insensitive matching. |
| `sequential` | Logical (default `TRUE`). `TRUE` = chained passes; `FALSE` = single combined scan. |
| `nthreads` | Positive integer per-call thread cap, or `NULL` for the RcppParallel default; `1` forces serial execution. |

**Returns:** character vector the same length as `x`.

```{r}
gsub_all(c("Mr\\.", "Mrs\\."), c("Mister", "Missus"), c("Mr. Smith", "Mrs. Jones"))
```

## String utilities

All four functions in this group preserve `names(x)` and `NA` handling
exactly like their base R equivalents.

### `ftrimws()`

```r
ftrimws(x, which = c("both", "left", "right"), whitespace = "[ \t\r\n]")
```

| Argument | Description |
|---|---|
| `x` | Character vector. `NA` elements return `NA`. |
| `which` | One of `"both"`, `"left"`, `"right"`. |
| `whitespace` | Regex of characters to strip. Only the default `"[ \t\r\n]"` uses the fast path; any other value delegates to `base::trimws()`. |

**Returns:** character vector the same length as `x`, with `names(x)`
preserved.

```{r}
ftrimws(c("  padded  ", "\ttabbed\n"))
ftrimws("  left only", which = "left")
```

### `fsubstr()`

```r
fsubstr(x, start, stop)
```

| Argument | Description |
|---|---|
| `x` | Character vector. `NA` elements return `NA`. |
| `start`, `stop` | Integer (or numeric) vectors of length 1 or `length(x)`, 1-indexed, clamped to each string's bounds, recycled to `length(x)`. `NA` in either produces `NA` for that element. |

**Returns:** character vector the same length as `x`.

```{r}
fsubstr("NSWREC123456789", 1, 3)
fsubstr(rep("ABCDEFGH", 3), c(1, 4, 7), c(3, 6, 8))  # vectorised start/stop
```

### `fnchar()`

```r
fnchar(x, type = "chars", allowNA = FALSE, keepNA = NA)
```

| Argument | Description |
|---|---|
| `x` | Vector, coerced to character via `as.character()` if needed. |
| `type` | One of `"bytes"`, `"chars"`, or `"width"`. `"width"` delegates to `base::nchar()` (display width isn't parallelised). |
| `allowNA` | Logical. Ignored; the fast path never raises an encoding error regardless. |
| `keepNA` | Logical or `NA` (default). `NA`/`TRUE`: `NA` elements of `x` return `NA`. `FALSE`: they return `2L` (length of the string `"NA"`), matching base R's legacy behaviour. |

**Returns:** integer vector the same length as `x`, with `names(x)`
preserved.

```{r}
fnchar(c("café", "abc", NA))
fnchar(c("café", "abc"), type = "bytes")  # "é" is 2 bytes in UTF-8
```

### `fchartr()`

```r
fchartr(old, new, x)
```

| Argument | Description |
|---|---|
| `old`, `new` | Single strings with the same number of characters. Each character of `old` maps to the corresponding character of `new`. |
| `x` | Character vector. `NA` elements return `NA`. |

**Returns:** character vector the same length as `x`. Falls back to
`base::chartr()` automatically when `old`/`new` contain multi-byte
characters (the fast path is a flat byte lookup table, valid only for
single-byte/ASCII characters).

```{r}
fchartr("'-", "  ", c("O'Brien", "Smith-Jones"))
fchartr("aeiou", "AEIOU", "fast string")
```

## Dates

### `format_date()`

```r
format_date(x, format = c("iso", "compact", "dmy", "ymd_slash"))
```

| Argument | Description |
|---|---|
| `x` | A `Date` object or numeric vector of days since 1970-01-01. |
| `format` | `"iso"` (`YYYY-MM-DD`), `"compact"` (`YYYYMMDD`), `"dmy"` (`DD/MM/YYYY`), or `"ymd_slash"` (`YYYY/MM/DD`). |

**Returns:** character vector the same length as `x`.

```{r}
format_date(as.Date(c("2024-06-18", NA)), "compact")
```

### `format_date_parts()`

Pure string-building (zero-padding plus punctuation, no calendar math, no
`Date` object) — the inverse of `date_parts()`.

```r
format_date_parts(year, month, day, format = c("iso", "compact", "dmy", "ymd_slash"))
```

| Argument | Description |
|---|---|
| `year`, `month`, `day` | Numeric vectors, recycled to a common length. `NA`, non-integer-coercible, or out-of-width values (year outside 0-9999, month/day outside 0-99) produce `NA` for that element. No calendar validation — `month = 13` formats as given if it fits the field width. |
| `format` | Same four options as `format_date()`. |

**Returns:** character vector recycled to the common length of `year`,
`month`, `day`.

```{r}
format_date_parts(2024, 6, 18, "iso")
```

### `date_parts()`

```r
date_parts(x)
```

| Argument | Description |
|---|---|
| `x` | A `Date` object or numeric vector of days since 1970-01-01. |

**Returns:** a `data.frame` with integer columns `year`, `month`, `day`.

```{r}
date_parts(as.Date(c("2024-06-18", "1999-12-31")))
```

### `fas.Date()`

Not a drop-in for `base::as.Date()` — see @sec-theory for the speed/
strictness trade-off (no days-in-month or leap-year validation).

```r
fas.Date(x, format = c("iso", "compact", "dmy", "ymd_slash"))
```

| Argument | Description |
|---|---|
| `x` | Character vector. `NA` elements, and elements that don't match `format` exactly (wrong length/separators/non-digits) or have an out-of-range month/day, become `NA`. |
| `format` | Same four options as `format_date()`; `x` must be in exactly this one format. |

**Returns:** a `Date` vector the same length as `x`.

```{r}
fas.Date(c("18/06/2024", "31/12/1999", "not-a-date"), "dmy")
```

## Phonetic blocking keys

All four functions have no base R equivalent and are intended as blocking
keys ahead of fuzzy name matching (@sec-theory, @sec-workflow).

### `soundex()`

```r
soundex(x)
```

| Argument | Description |
|---|---|
| `x` | Character vector (coerced via `as.character()` if not already character — e.g. a factor read from CSV). `NA` elements, and elements with no alphabetic characters, return `NA`. |

**Returns:** character vector the same length as `x`, with `names(x)`
preserved; each element either `NA` or exactly 4 characters.

```{r}
soundex(c("Robert", "Rupert", "Ashcraft"))
```

### `nysiis()`

```r
nysiis(x)
```

| Argument | Description |
|---|---|
| `x` | Character vector (coerced via `as.character()` if needed). `NA` elements, and elements with no alphabetic characters, return `NA`. |

**Returns:** character vector the same length as `x`, with `names(x)`
preserved; each element either `NA` or up to 6 characters.

```{r}
nysiis(c("Robert", "Rupert", "Ashcraft"))
```

### `double_metaphone()`

Two codes per word — `primary` and `secondary` — for ambiguous
pronunciations. Ported from the structure of Apache Commons Codec's
Java implementation and cross-checked against its published test
vectors. Unlike `soundex()`/`nysiis()`, internal whitespace/punctuation
is *not* stripped before encoding (only the ends are trimmed and the
text uppercased) — the algorithm itself keys off literal fragments like
`"VAN "`/`"SCH"`, including the space.

```r
double_metaphone(x)
```

| Argument | Description |
|---|---|
| `x` | Character vector (coerced via `as.character()` if needed). `NA` elements, and elements with no content after trimming, return `NA` in both columns. |

**Returns:** a `data.frame` with character columns `primary` and
`secondary`, each `length(x)` long.

```{r}
double_metaphone(c("Smith", "Schmidt", "Catherine", "Kathryn"))
```

### `caverphone()`

Caverphone 2.0 (Caversham Project, University of Otago) — a fixed
10-character code via a chain of substring transforms, widely used in
NZ/AU record linkage. Ported step-for-step from Apache Commons Codec's
`Caverphone2` and cross-checked against its published test vectors,
including the "enough"/"trough" duplicate-rule quirk inherited from the
original 2004 specification (preserved rather than "corrected" — the
point of a phonetic key is to match the standard everyone else uses).

```r
caverphone(x)
```

| Argument | Description |
|---|---|
| `x` | Character vector (coerced via `as.character()` if needed). `NA` elements return `NA`; every other input — including `""` — returns a 10-character code. |

**Returns:** character vector the same length as `x`, with `names(x)`
preserved; each element either `NA` or exactly 10 characters.

```{r}
caverphone(c("Peter", "Tedder", "Stevenson"))
```

## Fuzzy string similarity

### `jaro_winkler()`

Vectorised pairwise Jaro-Winkler similarity. See @sec-theory for the
algorithm and the bit-parallel implementation.

```r
jaro_winkler(a, b, p = 0.1, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `a`, `b` | Equal-length character vectors. |
| `p` | Prefix scaling factor (default `0.1`, the standard value). |
| `nthreads` | Positive integer per-call thread cap, or `NULL` for the RcppParallel default; `1` forces serial execution. |

**Returns:** numeric vector of similarities in `[0, 1]`.

```{r}
jaro_winkler(c("MARTHA", "DWAYNE"), c("MARHTA", "DUANE"))
jaro_winkler("SMITH", "SMYTH", p = 0)    # no prefix bonus
jaro_winkler("SMITH", "SMYTH", p = 0.1)  # standard prefix bonus
```

### `jaro_winkler_matrix()`

All-pairs Jaro-Winkler similarity — every element of `a` against every
element of `b`.

```r
jaro_winkler_matrix(a, b, p = 0.1, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `a` | Character vector of length n (rows). |
| `b` | Character vector of length m (columns). |
| `p`, `nthreads` | Same as `jaro_winkler()`. |

**Returns:** numeric matrix of dimensions n × m.

```{r}
jaro_winkler_matrix(c("JOHN SMITH", "MARY JONES"), c("JON SMYTH", "MARY JONES", "JOHN SMITH"))
```

### `jaro_winkler_tokens()`

Token-reorder- and punctuation-aware Jaro-Winkler similarity. Full
mechanics in @sec-theory.

```r
jaro_winkler_tokens(a, b, p = 0.1, ignore_case = FALSE, strip = "['’-]",
                     extra_penalty = NULL, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `a`, `b` | Equal-length character vectors. |
| `p` | Prefix scaling factor (default `0.1`). |
| `ignore_case` | Logical (default `FALSE`). Uppercase both sides before comparing. |
| `strip` | Single regex string matching punctuation to remove before tokenising, or `NULL` to skip. Default `"['’-]"` strips straight/curly apostrophes and hyphens. |
| `extra_penalty` | `NULL` (default) or a non-negative numeric scalar. `NULL`: unmatched tokens dilute the average (divide by `max(n_tokens_a, n_tokens_b)`). A number: average over only matched tokens, then subtract `extra_penalty` per unmatched token (floored at 0) — `0` ignores extra tokens outright. |
| `nthreads` | Positive integer per-call thread cap, or `NULL` for the RcppParallel default; `1` forces serial execution. |

**Returns:** numeric vector of similarities in `[0, 1]`. `NA` if either
side of a pair is `NA`.

```{r}
# Reordered tokens, one near-miss — rescued relative to plain jaro_winkler()
jaro_winkler_tokens("Kyle John Haynes", "John Kylie Haynes")
jaro_winkler("Kyle John Haynes", "John Kylie Haynes")

# Punctuation/spacing variants of the same name
jaro_winkler_tokens(c("OBrien", "OBrien"), c("O Brien", "O'Brien"))

# A stray extra token, penalised by default...
jaro_winkler_tokens("Kylie John ZZ Haynes", "Haynes John Kyle")
# ...vs. ignored outright with extra_penalty = 0
jaro_winkler_tokens("Kylie John ZZ Haynes", "Haynes John Kyle", extra_penalty = 0)
# ...vs. lightly discounted
jaro_winkler_tokens("Kylie John ZZ Haynes", "Haynes John Kyle", extra_penalty = 0.1)
```

## Edit distance

`levenshtein()` and `damerau_levenshtein()` use a bit-parallel
implementation (Myers' (1999) bit-vector algorithm for plain Levenshtein,
a three-rolling-row dynamic program for Damerau-Levenshtein) rather than
`stringdist`'s generic dispatch — see @sec-theory for the algorithm and
@sec-benchmarks for the speed comparison.

### `levenshtein()`

```r
levenshtein(a, b, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `a`, `b` | Equal-length character vectors. |
| `nthreads` | Positive integer per-call thread cap, or `NULL` for the RcppParallel default; `1` forces serial execution. |

**Returns:** numeric vector the same length as `a` (minimum number of
single-character insertions/deletions/substitutions). `NA` if either side
of a pair is `NA`.

```{r}
levenshtein("kitten", "sitting")
```

### `levenshtein_matrix()`

All-pairs Levenshtein distance.

```r
levenshtein_matrix(a, b, nthreads = NULL)
```

**Returns:** numeric matrix of dimensions `length(a)` x `length(b)`.

```{r}
levenshtein_matrix(c("kitten", "cat"), c("sitting", "bat"))
```

### `damerau_levenshtein()`

Like `levenshtein()`, but an adjacent-character transposition
(`"ab"` <-> `"ba"`) also costs one edit, under the "optimal string
alignment" (OSA) restriction that no substring is edited more than once —
matching `stringdist::stringdist(method = "dl")`, not full unrestricted
Damerau-Levenshtein.

```r
damerau_levenshtein(a, b, nthreads = NULL)
```

Arguments identical to `levenshtein()`. **Returns:** numeric vector the
same length as `a`.

```{r}
damerau_levenshtein("ab", "ba")   # one transposition
levenshtein("ab", "ba")           # two substitutions
```

### `damerau_levenshtein_matrix()`

All-pairs Damerau-Levenshtein (OSA) distance — same shape as
`levenshtein_matrix()`.

```r
damerau_levenshtein_matrix(a, b, nthreads = NULL)
```

### `hamming()`

Number of positions at which two equal-length strings differ.

```r
hamming(a, b, nthreads = NULL)
```

Arguments identical to `levenshtein()`. **Returns:** numeric vector the
same length as `a`; `Inf` for a pair of unequal length (matching
`stringdist::stringdist(method = "hamming")`).

```{r}
hamming("karolin", "kathrin")
hamming("abc", "ab")  # Inf -- unequal length
```

## Q-gram set similarity

`jaccard_index()`, `dice_coefficient()`, and `tversky_index()` compare two
strings' sets of overlapping length-`q` substrings ("q-grams"). All three
share one sorted-merge intersection pass over packed-integer q-gram sets
without routing through `stringdist`'s generic R-level dispatch
(@sec-benchmarks).

### `jaccard_index()`

```r
jaccard_index(a, b, q = 2, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `a`, `b` | Equal-length character vectors. |
| `q` | Q-gram length (default `2`, i.e. bigrams). Must be a single integer >= 1. |
| `nthreads` | Positive integer per-call thread cap, or `NULL` for the RcppParallel default; `1` forces serial execution. |

**Returns:** numeric vector of similarities in `[0, 1]`, `length(a)`
long. Two strings shorter than `q` (so neither has any q-grams) compare
equal (`1`).

```{r}
jaccard_index("night", "nacht")
```

### `jaccard_matrix()`

All-pairs Jaccard similarity. Same argument/return shape as
`jaro_winkler_matrix()`.

```r
jaccard_matrix(a, b, q = 2, nthreads = NULL)
```

### `dice_coefficient()` / `dice_matrix()`

The Sorensen-Dice coefficient: `2*|intersection| / (|A| + |B|)`. Weights
shared q-grams more heavily than Jaccard (always `>=` the Jaccard score
for the same pair). Same arguments as `jaccard_index()`/`jaccard_matrix()`.

```{r}
dice_coefficient("night", "nacht")
```

### `tversky_index()` / `tversky_matrix()`

Generalises both: `|intersection| / (|intersection| + alpha*|A only| +
beta*|B only|)`. `alpha = beta = 1` reduces to Jaccard; `alpha = beta =
0.5` (the default) reduces to Dice.

```r
tversky_index(a, b, q = 2, alpha = 0.5, beta = 0.5, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `a`, `b`, `q`, `nthreads` | Same as `jaccard_index()`. |
| `alpha`, `beta` | Tversky asymmetry weights (default `0.5` each). Non-negative scalars. |

```{r}
tversky_index("night", "nacht", alpha = 1, beta = 1)   # == jaccard_index()
tversky_index("night", "nacht", alpha = 0.5, beta = 0.5) # == dice_coefficient()
```

## fuzzywuzzy-style ratios

Vectorised ports of Python's
[fuzzywuzzy](https://github.com/seatgeek/fuzzywuzzy) package, implementing
the same Ratcliff/Obershelp matching-blocks algorithm fuzzywuzzy itself
falls back to (the one behind Python's `difflib.SequenceMatcher`). All
four return `0-100` like the Python originals; see @sec-theory for the
algorithm.

```r
fuzz_ratio(a, b, full_process = TRUE, nthreads = NULL)
fuzz_partial_ratio(a, b, full_process = TRUE, nthreads = NULL)
fuzz_token_sort_ratio(a, b, full_process = TRUE, nthreads = NULL)
fuzz_token_set_ratio(a, b, full_process = TRUE, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `a`, `b` | Equal-length character vectors. |
| `full_process` | Logical (default `TRUE`, matching fuzzywuzzy's default). Lowercases and replaces runs of non-alphanumeric characters with a single space before comparing. |
| `nthreads` | Positive integer per-call thread cap, or `NULL` for the RcppParallel default; `1` forces serial execution. |

**Returns:** numeric vector of scores in `[0, 100]`, `length(a)` long.

- **`fuzz_ratio()`** — overall similarity: `2*M / (len(a)+len(b))`, `M`
  being the total length of the longest-common matching blocks (found
  recursively — *not* edit distance).
- **`fuzz_partial_ratio()`** — best alignment of the shorter string against
  any equal-length window of the longer one.
- **`fuzz_token_sort_ratio()`** — splits into whitespace tokens, sorts
  them, rejoins, then runs `fuzz_ratio()` — word order stops mattering.
- **`fuzz_token_set_ratio()`** — splits into token *sets* and compares the
  shared-token core against each side's leftovers, taking the best of
  three pairwise ratios — robust to one side having extra words.

```{r}
fuzz_ratio("this is a test", "this is a test!")
fuzz_partial_ratio("fuzzy wuzzy was a bear", "wuzzy fuzzy was a bear")
fuzz_token_sort_ratio("fuzzy was a bear", "bear was a fuzzy")
fuzz_token_set_ratio("fuzzy was a bear", "fuzzy fuzzy bear was a bear")
```
