# Preface {.unnumbered}

`fast.string` is an R package of drop-in-faster equivalents for the string,
date, and phonetic-matching functions that show up constantly in record
linkage and ETL work — `grepl()`, `grep()`, `sub()`, `gsub()`, `trimws()`,
`substr()`, `nchar()`, `chartr()`, plus a handful of functions base R simply
doesn't have at all: Soundex/NYSIIS phonetic codes, and vectorised
Jaro-Winkler similarity (including a token-reorder/punctuation-aware
variant for multi-word names).

Large workloads are parallelised via Intel TBB (through
[RcppParallel](https://rcppcore.github.io/RcppParallel/)), with
[PCRE2](https://www.pcre.org/) doing the regex work and
a prepared byte-oriented searcher handling fixed/literal matching.
Performance depends on string length, match density, value reuse, matrix
shape, and available cores. Some operations (parallel Jaro-Winkler,
multi-pattern substitution in one pass) have no base R equivalent at all.

## Why this exists

Record linkage and data-cleaning pipelines spend a surprising fraction of
their wall-clock time in string operations that look trivial in isolation —
`trimws()` on a column, a `gsub()` to strip punctuation, `grepl()` to flag a
pattern — but get called over and over on multi-million-row name/address
files. Individually each call is fast; in aggregate, on a 5M-row file, they
add up to minutes of CPU time spent doing work a single core could be doing
across eight or sixteen instead.

Three things are conspicuously *not* one function each in base R, despite
being everywhere in this kind of pipeline:

- **Phonetic blocking keys** (Soundex, NYSIIS) — used to group
  *probably-same* names before running an expensive pairwise comparison
  within each group, rather than comparing every record against every
  other record.
- **Fuzzy string similarity** (Jaro-Winkler) — base R has no notion of
  "how similar are these two strings," only exact (`==`) or pattern
  (`grepl()`) matching.
- **Multi-token name comparison** — `"Kyle John Haynes"` and
  `"John Kylie Haynes"` are obviously the same person with two tokens
  swapped, but a plain character-by-character comparison scores that badly.

`fast.string` exists to make all of the above fast enough to run
unblinkingly on real-sized data, and to package the boring-but-essential
80% (trimming, substitution, date parsing) so a project doesn't need five
different dependencies for five different speed problems.

## How it fits with record-linkage packages

`fast.string` is a toolbox of vectorised string primitives. The
blocking-strategy selection, EM-based match probability estimation, and
persistence of a linked dataset are handled well already by packages like
[`RecordLinkage`](https://cran.r-project.org/package=RecordLinkage) or
[`reclin2`](https://cran.r-project.org/package=reclin2) — pair `fast.string`
with one of those for the per-string comparison work they call underneath.
@sec-benchmarks compares `fast.string`'s comparison primitives directly
against the equivalent functions in `RecordLinkage` and `stringdist` at
scale.

## How this book is organised

- **Getting started** — what `fast.string` is for and how it's positioned
  against base R, `stringi`/`stringr`, `stringdist`, and `RecordLinkage`
  (@sec-about), then a hands-on tour of every function group
  (@sec-quickstart).
- **Under the hood** — the algorithms (@sec-theory): how regex and
  fixed-literal matching differ, how the bit-parallel Jaro-Winkler
  implementation works, and how multi-token name comparison resolves
  reordered tokens — followed by a
  full worked record-linkage pipeline (@sec-workflow) putting the pieces
  together.
- **Reference** — an exhaustive per-function reference with every
  argument and runnable examples (@sec-reference); the heavy benchmark
  chapter at **3 million strings**, with `base`, `stringi`, `stringdist`,
  and `RecordLinkage` as comparison baselines (@sec-benchmarks); and a
  troubleshooting/FAQ chapter (@sec-troubleshooting).

## A quick taste

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

```{r}
#| label: teaser
n <- 2e5
x <- sprintf("%s-%04d", sample(c("John Smith", "Mary O'Brien", "Kyle Haynes"), n, TRUE), sample(9999, n, TRUE))

system.time(base::grepl("Smith", x))
system.time(fast.string::fgrepl("Smith", x))

round(fast.string::jaro_winkler_tokens("Kyle John Haynes", "John Kylie Haynes"), 3)
```

That last call is the headline example for the multi-token comparison
function: two names with reordered tokens and one typo still score above
0.95. The full story — including why it also handles `"O'Brien"` vs.
`"O Brien"` vs. `"OBrien"` — is in @sec-theory and @sec-reference.
