---
title: "fast.string: Parallel String, Date & Phonetic Matching"
subtitle: "PCRE2 + prepared literal search + RcppParallel (Intel TBB)"
author: "Kyle Haynes"
date: today
format:
  html:
    toc: true
    code-fold: show
    df-print: paged
execute:
  warning: false
---

## Overview

`fast.string` provides fast equivalents of common base R string functions
(`grepl()`, `grep()`, `sub()`, `gsub()`, `trimws()`, `substr()`, `nchar()`,
`chartr()`), plus date formatting/parsing and phonetic-matching helpers
(`soundex()`, `nysiis()`, `jaro_winkler()`), with large workloads
parallelised across CPU cores via Intel TBB (through RcppParallel), PCRE2
for regex, and a prepared byte-oriented searcher for fixed-string matching.

Performance depends on string length, match density, value reuse, matrix
shape, and available cores. The benchmark chapter measures those factors
against an isolated installation of the exact checkout being documented.

```{r}
#| label: setup
#| include: false
source("benchmark_helpers.R", local = FALSE)
.benchmark_installation <- benchmark_install_checkout(".")
.benchmark_metadata <- benchmark_metadata(
    .benchmark_installation,
    profile = "fast-string-qmd"
)
options(fast.string.verbose = FALSE)
library(stringi)
```

All examples below time a single run with `system.time()` against a
moderately-sized vector (200,000 strings, lengths 2–40, with a few `NA`
and `""` mixed in) so this document renders quickly — see
[`Benchmarks.R`](Benchmarks.R) in the repo for the full 2M-row,
`microbenchmark`-based benchmark suite.

```{r}
#| label: test-data
set.seed(1)
n <- 2e5
x <- stri_rand_strings(n, sample(2:40, n, replace = TRUE), pattern = "[A-Za-z0-9 ]")
na_idx    <- sample(n, 200)
empty_idx <- sample(setdiff(seq_len(n), na_idx), 200)
x[na_idx]    <- NA_character_
x[empty_idx] <- ""

length(x)
```

## Matching & substitution

`fgrepl()`, `fgrep()`, `fsub()`, and `fgsub()` mirror `grepl()`, `grep()`,
`sub()`, and `gsub()` exactly (same arguments, same `NA`/edge-case
behaviour), but run the matching/substitution in parallel.

```{r}
#| label: grepl-bench
system.time(base::grepl("[0-9]{2}", x))
system.time(fast.string::fgrepl("[0-9]{2}", x))
```

```{r}
#| label: gsub-bench
system.time(base::gsub("[0-9]+", "#", x, perl = TRUE))
system.time(fast.string::fgsub("[0-9]+", "#", x))
```

`fcount()` counts non-overlapping matches per element — the counting
counterpart to `fgrepl()`, replacing the `vapply(gregexpr(...), length, ...)`
idiom that allocates a vector of match positions per row just to take its
length.

```{r}
#| label: count-bench
system.time(vapply(base::gregexpr("[0-9]", x),
                   function(m) if (m[1] < 0L) 0L else length(m), integer(1)))
system.time(fast.string::fcount("[0-9]", x))
```

`gsub_all()` has no base equivalent: it applies several patterns in a
single pass, either sequentially (chaining, like repeated `gsub()` calls)
or as one combined left-to-right scan.

```{r}
#| label: gsub-all-bench
patterns <- c("a", "e", "i", "o", "u")
repls    <- c("4", "3", "1", "0", "*")
system.time({
    y <- x
    for (i in seq_along(patterns)) y <- base::gsub(patterns[i], repls[i], y, fixed = TRUE)
})
system.time(fast.string::gsub_all(patterns, repls, x, fixed = TRUE, sequential = TRUE))
system.time(fast.string::gsub_all(patterns, repls, x, fixed = TRUE, sequential = FALSE))
```

## String utilities

`ftrimws()`, `fsubstr()`, `fnchar()`, and `fchartr()` mirror their base
equivalents, preserving `names(x)` and `NA` handling exactly.

```{r}
#| label: str-utils-bench
x_ws <- ifelse(is.na(x), NA_character_, paste0("  ", x, "\t\n"))

system.time(base::trimws(x_ws))
system.time(fast.string::ftrimws(x_ws))

system.time(base::nchar(x, "chars"))
system.time(fast.string::fnchar(x, "chars"))
```

## Dates

`format_date()` / `fas.Date()` trade `base::as.Date()`/`format.Date()`'s
locale-aware, multi-format flexibility for raw speed on one fixed format
at a time. `date_parts()` decomposes a `Date` vector into year/month/day
columns in a single pass — handy for blocking keys in record linkage.

```{r}
#| label: dates-bench
dates <- as.Date(sample(seq(as.Date("1950-01-01"), as.Date("2024-12-31"), by = "day"), n, replace = TRUE))
dates[na_idx] <- NA

system.time(format(dates, "%Y-%m-%d"))
system.time(fast.string::format_date(dates, "iso"))
```

```{r}
#| label: date-parts-demo
head(fast.string::date_parts(dates))
```

`fas.POSIXct()` and `format_datetime()` do the same for timestamps,
working entirely in UTC seconds since the epoch — no timezone-database
lookup. Unlike `fas.Date()`, they validate the full calendar (month
lengths and leap years) and return `NA` for impossible values.

```{r}
#| label: datetime-bench
stamps <- format(as.POSIXct("2024-01-01", tz = "UTC") + sample(3e7, n), "%Y-%m-%d %H:%M:%S")

system.time(base::as.POSIXct(stamps, tz = "UTC"))
system.time(fast.string::fas.POSIXct(stamps))
```

```{r}
#| label: datetime-demo
ts <- fast.string::fas.POSIXct(c("2024-06-18 09:15:00", "2023-02-29 00:00:00"))
ts
fast.string::format_datetime(ts, "iso_offset", offset = "+10:00")
```

## Phonetic blocking keys

`soundex()`, `nysiis()`, `refined_soundex()`, and `cologne()` have no base
R equivalent. They're typically used as blocking keys ahead of fuzzy name
matching, grouping records that are *phonetically* similar before running
a more expensive comparison like Jaro-Winkler within each block.

```{r}
#| label: phonetic-demo
names_x <- c("Robert", "Rupert", "Ashcraft", "Ashcroft", NA)
data.frame(name = names_x, soundex = fast.string::soundex(names_x), nysiis = fast.string::nysiis(names_x))
```

The schemes differ in how coarsely they group. `refined_soundex()` encodes
vowels too and doesn't truncate, giving smaller blocks; `cologne()` folds
German umlauts to their base vowels, so spelling variants that ASCII-only
schemes split are grouped together.

```{r}
#| label: phonetic-extended-demo
german <- c("Müller", "Mueller", "Miller", "Schmidt", "Schmitt")
data.frame(
    name    = german,
    soundex = fast.string::soundex(german),
    refined = fast.string::refined_soundex(german),
    cologne = fast.string::cologne(german)
)
```

## Fuzzy string similarity

`jaro_winkler()` compares two equal-length vectors pairwise;
`jaro_winkler_matrix()` computes the full n × m similarity matrix between
two vectors — the typical shape of a data-linkage blocking table.

```{r}
#| label: jaro-bench
system.time(fast.string::jaro_winkler(x, rev(x)))
```

```{r}
#| label: jaro-matrix-demo
candidates_a <- c("John Smith", "Mary Jones", "Robert Brown")
candidates_b <- c("Jon Smith", "Mary-Anne Jones", "Rob Brown", "Someone Else")
fast.string::jaro_winkler_matrix(candidates_a, candidates_b)
```

A second, independent similarity signal is often worth carrying alongside
Jaro-Winkler. `cosine_similarity()` compares q-gram *frequency* profiles,
so unlike the set-based `jaccard_index()`/`dice_coefficient()`/
`tversky_index()` it keeps repeated q-grams — and unlike
`jaro_winkler_tokens()` it is sensitive to token order.

```{r}
#| label: cosine-demo
a <- c("aaaa", "KYLE JOHN HAYNES", "night")
b <- c("aaab", "HAYNES JOHN KYLE", "nacht")
data.frame(a, b,
           jaccard = fast.string::jaccard_index(a, b),
           cosine  = fast.string::cosine_similarity(a, b))
```

## Further reading

- [`README.md`](README.md) — quick install and `system.time()` examples.
- [The documentation book](https://kylehaynes.github.io/fast.string/) — a
  narrative walkthrough of every function group, the algorithms behind
  them, and a full worked record-linkage pipeline.
- [`Benchmarks.R`](Benchmarks.R) — full benchmark suite at 2M-row scale,
  covering every exported function.
- [`examples_data_linkage.R`](examples_data_linkage.R) — an end-to-end
  record-linkage pipeline using these functions together.
