---
title: "fast.string: Parallel String, Date & Phonetic Matching"
subtitle: "PCRE2 + RE2 + RcppParallel (Intel TBB)"
author: "Kyle Haynes"
date: today
format:
  html:
    toc: true
    code-fold: show
    df-print: paged
execute:
  warning: false
---

## Overview

`fast.string` provides fast equivalents of common base R string functions
(`grepl()`, `grep()`, `sub()`, `gsub()`, `trimws()`, `substr()`, `nchar()`,
`chartr()`), plus date formatting/parsing and phonetic-matching helpers
(`soundex()`, `nysiis()`, `jaro_winkler()`) — all parallelised across CPU
cores via Intel TBB (through RcppParallel), with PCRE2 for regex and RE2
`StringPiece` for fixed-string matching.

These are typically 5–40x faster than the base R equivalents on large
character vectors, which matters most for record-linkage and ETL pipelines
that process millions of strings.

```{r}
#| label: setup
#| include: false
if (!requireNamespace("fast.string", quietly = TRUE)) {
    remotes::install_local(".", quiet = TRUE, upgrade = "never")
}
library(fast.string)
options(fast.string.verbose = FALSE)
library(stringi)
```

All examples below time a single run with `system.time()` against a
moderately-sized vector (200,000 strings, lengths 2–40, with a few `NA`
and `""` mixed in) so this document renders quickly — see
[`Benchmarks.R`](Benchmarks.R) in the repo for the full 2M-row,
`microbenchmark`-based benchmark suite.

```{r}
#| label: test-data
set.seed(1)
n <- 2e5
x <- stri_rand_strings(n, sample(2:40, n, replace = TRUE), pattern = "[A-Za-z0-9 ]")
na_idx    <- sample(n, 200)
empty_idx <- sample(setdiff(seq_len(n), na_idx), 200)
x[na_idx]    <- NA_character_
x[empty_idx] <- ""

length(x)
```

## Matching & substitution

`fgrepl()`, `fgrep()`, `fsub()`, and `fgsub()` mirror `grepl()`, `grep()`,
`sub()`, and `gsub()` exactly (same arguments, same `NA`/edge-case
behaviour), but run the matching/substitution in parallel.

```{r}
#| label: grepl-bench
system.time(base::grepl("[0-9]{2}", x))
system.time(fast.string::fgrepl("[0-9]{2}", x))
```

```{r}
#| label: gsub-bench
system.time(base::gsub("[0-9]+", "#", x, perl = TRUE))
system.time(fast.string::fgsub("[0-9]+", "#", x))
```

`gsub_all()` has no base equivalent: it applies several patterns in a
single pass, either sequentially (chaining, like repeated `gsub()` calls)
or as one combined left-to-right scan.

```{r}
#| label: gsub-all-bench
patterns <- c("a", "e", "i", "o", "u")
repls    <- c("4", "3", "1", "0", "*")
system.time({
    y <- x
    for (i in seq_along(patterns)) y <- base::gsub(patterns[i], repls[i], y, fixed = TRUE)
})
system.time(fast.string::gsub_all(patterns, repls, x, fixed = TRUE, sequential = TRUE))
system.time(fast.string::gsub_all(patterns, repls, x, fixed = TRUE, sequential = FALSE))
```

## String utilities

`ftrimws()`, `fsubstr()`, `fnchar()`, and `fchartr()` mirror their base
equivalents, preserving `names(x)` and `NA` handling exactly.

```{r}
#| label: str-utils-bench
x_ws <- ifelse(is.na(x), NA_character_, paste0("  ", x, "\t\n"))

system.time(base::trimws(x_ws))
system.time(fast.string::ftrimws(x_ws))

system.time(base::nchar(x, "chars"))
system.time(fast.string::fnchar(x, "chars"))
```

## Dates

`format_date()` / `fas.Date()` trade `base::as.Date()`/`format.Date()`'s
locale-aware, multi-format flexibility for raw speed on one fixed format
at a time. `date_parts()` decomposes a `Date` vector into year/month/day
columns in a single pass — handy for blocking keys in record linkage.

```{r}
#| label: dates-bench
dates <- as.Date(sample(seq(as.Date("1950-01-01"), as.Date("2024-12-31"), by = "day"), n, replace = TRUE))
dates[na_idx] <- NA

system.time(format(dates, "%Y-%m-%d"))
system.time(fast.string::format_date(dates, "iso"))
```

```{r}
#| label: date-parts-demo
head(fast.string::date_parts(dates))
```

## Phonetic blocking keys

`soundex()` and `nysiis()` have no base R equivalent. They're typically
used as blocking keys ahead of fuzzy name matching, grouping records that
are *phonetically* similar before running a more expensive comparison
like Jaro-Winkler within each block.

```{r}
#| label: phonetic-demo
names_x <- c("Robert", "Rupert", "Ashcraft", "Ashcroft", NA)
data.frame(name = names_x, soundex = fast.string::soundex(names_x), nysiis = fast.string::nysiis(names_x))
```

## Fuzzy string similarity

`jaro_winkler()` compares two equal-length vectors pairwise;
`jaro_winkler_matrix()` computes the full n × m similarity matrix between
two vectors — the typical shape of a data-linkage blocking table.

```{r}
#| label: jaro-bench
system.time(fast.string::jaro_winkler(x, rev(x)))
```

```{r}
#| label: jaro-matrix-demo
candidates_a <- c("John Smith", "Mary Jones", "Robert Brown")
candidates_b <- c("Jon Smith", "Mary-Anne Jones", "Rob Brown", "Someone Else")
fast.string::jaro_winkler_matrix(candidates_a, candidates_b)
```

## Further reading

- [`README.md`](README.md) — quick install and `system.time()` examples.
- [`Benchmarks.R`](Benchmarks.R) — full benchmark suite at 2M-row scale,
  covering every exported function.
- [`examples_data_linkage.R`](examples_data_linkage.R) — an end-to-end
  record-linkage pipeline using these functions together.
