# Quickstart {#sec-quickstart}

```{r}
#| label: setup
#| include: false
source("_setup.R")
suppressPackageStartupMessages(library(stringi))
```

## Installation

```r
remotes::install_github("KyleHaynes/anthropic_fast")
```

This compiles C++ source on install, so you need a working toolchain:
Rtools on Windows, Xcode command-line tools on macOS, or `build-essential`
+ `libpcre2-dev` on Linux. See @sec-troubleshooting if the install fails.

```{r}
#| label: load
library(fast.string)
```

Loading the package prints a one-time startup banner listing every
exported function; suppress it with `options(fast.string.verbose = FALSE)`
set *before* `library()`, or with `suppressPackageStartupMessages()`.

The rest of this chapter is a guided tour through every function group,
using small inputs so it renders fast — @sec-benchmarks is where the same
functions get pushed to 3 million rows.

## Matching & substitution

`fgrepl()`, `fgrep()`, `fsub()`, and `fgsub()` are argument-compatible with
`base::grepl()`/`grep()`/`sub()`/`gsub()` — same defaults, same `NA`
handling — just parallelised.

```{r}
#| label: matching
x <- c("invoice-2024-001", "invoice-2024-002", "credit-note-001", NA)

fgrepl("^invoice", x)
fgrep("^invoice", x, value = TRUE)
fsub("-([0-9]+)$", "_\\1", x)
fgsub("[0-9]", "#", x)
```

`gsub_all()` has no base R equivalent: it applies several patterns in one
pass, either sequentially (chained, like calling `gsub()` repeatedly) or as
a single combined left-to-right scan where the first matching pattern at
each position wins.

```{r}
#| label: gsub-all
patterns <- c("Mr\\.", "Mrs\\.", "Dr\\.")
repls    <- c("Mister", "Missus", "Doctor")
gsub_all(patterns, repls, c("Mr. Smith", "Dr. Jones"), fixed = FALSE)
```

## String utilities

`ftrimws()`, `fsubstr()`, `fnchar()`, and `fchartr()` mirror their base
equivalents exactly, including `names(x)` preservation.

```{r}
#| label: str-utils
messy <- c("  John Smith  ", "\tMary Jones\n", NA)
ftrimws(messy)

records <- c("NSWREC123456789", "VICPOL987654321")
fsubstr(records, 1, 3)   # state code
fsubstr(records, 4, 6)   # category code

fnchar(c("café", NA, ""), type = "chars")

fchartr("'-", "  ", c("O'Brien", "Smith-Jones"))
```

## Dates

`format_date()`/`fas.Date()` trade `base::as.Date()`'s locale-aware,
multi-format flexibility for speed on one fixed format at a time.
`date_parts()` decomposes a `Date` vector into year/month/day columns in a
single pass — useful for building blocking keys.

```{r}
#| label: dates
dates <- as.Date(c("2024-06-18", "1999-12-31", NA))
format_date(dates, "compact")
date_parts(dates)
fas.Date(c("18/06/2024", "31/12/1999"), "dmy")
```

## Phonetic blocking keys

`soundex()` and `nysiis()` have no base R equivalent. They're typically
used to group records that *sound* alike before running a more expensive
pairwise comparison only within each group, rather than comparing every
record against every other record.

```{r}
#| label: phonetic
names_x <- c("Robert", "Rupert", "Ashcraft", "Ashcroft", "Smith", "Smyth")
data.frame(
    name    = names_x,
    soundex = soundex(names_x),
    nysiis  = nysiis(names_x)
)
```

`"Ashcraft"`/`"Ashcroft"` sharing a Soundex code (`A261`) despite differing
spelling is the textbook example of why Soundex exists — and a known point
of disagreement between *implementations* of Soundex, covered in
@sec-theory.

## Fuzzy string similarity

Three functions, three shapes of comparison:

- **`jaro_winkler(a, b)`** — pairwise, equal-length vectors. `a[i]` vs.
  `b[i]` for every `i`.
- **`jaro_winkler_matrix(a, b)`** — all-pairs. Every element of `a` against
  every element of `b`, returned as an `n × m` matrix.
- **`jaro_winkler_tokens(a, b)`** — pairwise like the first, but tolerant
  of token reordering and punctuation/spacing differences within a token.

```{r}
#| label: jaro
jaro_winkler(c("SMITH", "DWAYNE"), c("SMYTH", "DUANE"))

candidates_a <- c("John Smith", "Mary Jones")
candidates_b <- c("Jon Smith", "Mary-Anne Jones", "Someone Else")
jaro_winkler_matrix(candidates_a, candidates_b)

# Token reordering — two of three tokens match exactly, the third is close
jaro_winkler_tokens("Kyle John Haynes", "John Kylie Haynes")

# Punctuation/spacing variants of the same name
jaro_winkler_tokens(c("OBrien", "OBrien"), c("O Brien", "O'Brien"))
```

The full mechanics of how `jaro_winkler_tokens()` resolves reordered
tokens — and the `extra_penalty` argument for handling a stray extra
token — are in @sec-theory and the function reference (@sec-reference).

## Putting it together

A typical pipeline normalises with `fchartr()`/`ftrimws()`, builds a
blocking key with `soundex()`/`nysiis()`/`format_date()`, then scores
candidates within each block with `jaro_winkler_matrix()` or
`jaro_winkler_tokens()`. @sec-workflow walks through exactly that, end to
end, on a small linkage example.
