# Quickstart {#sec-quickstart}

```{r}
#| label: setup
#| include: false
source("_setup.R")
suppressPackageStartupMessages(library(stringi))
```

## Installation

```r
remotes::install_github("KyleHaynes/fast.string")
```

This compiles C++ source on install, so you need a working toolchain:
Rtools on Windows, Xcode command-line tools on macOS, or `build-essential`
+ `libpcre2-dev` on Linux. See @sec-troubleshooting if the install fails.

```{r}
#| label: load
library(fast.string)
```

Loading the package prints a one-time startup banner listing every
exported function; suppress it with `options(fast.string.verbose = FALSE)`
set *before* `library()`, or with `suppressPackageStartupMessages()`.

The rest of this chapter is a guided tour through every function group,
using small inputs so it renders fast — @sec-benchmarks is where the same
functions get pushed to 3 million rows.

## Matching & substitution

`fgrepl()`, `fgrep()`, `fsub()`, and `fgsub()` are argument-compatible with
`base::grepl()`/`grep()`/`sub()`/`gsub()` — same defaults, same `NA`
handling — just parallelised.

```{r}
#| label: matching
x <- c("invoice-2024-001", "invoice-2024-002", "credit-note-001", NA)

fgrepl("^invoice", x)
fgrep("^invoice", x, value = TRUE)
fsub("-([0-9]+)$", "_\\1", x)
fgsub("[0-9]", "#", x)
```

`gsub_all()` has no base R equivalent: it applies several patterns in one
pass, either sequentially (chained, like calling `gsub()` repeatedly) or as
a single combined left-to-right scan where the first matching pattern at
each position wins.

```{r}
#| label: gsub-all
patterns <- c("Mr\\.", "Mrs\\.", "Dr\\.")
repls    <- c("Mister", "Missus", "Doctor")
gsub_all(patterns, repls, c("Mr. Smith", "Dr. Jones"), fixed = FALSE)
```

## String utilities

`ftrimws()`, `fsubstr()`, `fnchar()`, and `fchartr()` mirror their base
equivalents exactly, including `names(x)` preservation.

```{r}
#| label: str-utils
messy <- c("  John Smith  ", "\tMary Jones\n", NA)
ftrimws(messy)

records <- c("NSWREC123456789", "VICPOL987654321")
fsubstr(records, 1, 3)   # state code
fsubstr(records, 4, 6)   # category code

fnchar(c("café", NA, ""), type = "chars")

fchartr("'-", "  ", c("O'Brien", "Smith-Jones"))
```

## Dates

`format_date()`/`fas.Date()` trade `base::as.Date()`'s locale-aware,
multi-format flexibility for speed on one fixed format at a time.
`date_parts()` decomposes a `Date` vector into year/month/day columns in a
single pass — useful for building blocking keys.

```{r}
#| label: dates
dates <- as.Date(c("2024-06-18", "1999-12-31", NA))
format_date(dates, "compact")
date_parts(dates)
fas.Date(c("18/06/2024", "31/12/1999"), "dmy")
```

## Phonetic blocking keys

`soundex()`, `nysiis()`, `double_metaphone()`, and `caverphone()` have no
base R equivalent. They're typically used to group records that *sound*
alike before running a more expensive pairwise comparison only within each
group, rather than comparing every record against every other record.
Different rulesets group different near-miss spellings together, so trying
more than one can catch matches a single phonetic key misses.

```{r}
#| label: phonetic
names_x <- c("Robert", "Rupert", "Ashcraft", "Ashcroft", "Smith", "Smyth")
data.frame(
    name       = names_x,
    soundex    = soundex(names_x),
    nysiis     = nysiis(names_x),
    caverphone = caverphone(names_x)
)
double_metaphone(c("Smith", "Schmidt", "Catherine", "Kathryn"))
```

`"Ashcraft"`/`"Ashcroft"` sharing a Soundex code (`A261`) despite differing
spelling is the textbook example of why Soundex exists — and a known point
of disagreement between *implementations* of Soundex, covered in
@sec-theory.

`double_metaphone()` returns two codes per word — `primary` and
`secondary` — because some spellings have more than one plausible
pronunciation (a `C` that could be a hard `K` or a soft `S` sound, for
instance); two names are usually considered a phonetic match if *any* of
one side's codes matches *any* of the other's. `"Catherine"` and
`"Kathryn"` sharing the code `K0RN` despite no letters lining up is the
kind of match Soundex/NYSIIS — designed around English/American surnames —
typically miss.

## Fuzzy string similarity

Three functions, three shapes of comparison:

- **`jaro_winkler(a, b)`** — pairwise, equal-length vectors. `a[i]` vs.
  `b[i]` for every `i`.
- **`jaro_winkler_matrix(a, b)`** — all-pairs. Every element of `a` against
  every element of `b`, returned as an `n × m` matrix.
- **`jaro_winkler_tokens(a, b)`** — pairwise like the first, but tolerant
  of token reordering and punctuation/spacing differences within a token.

```{r}
#| label: jaro
jaro_winkler(c("SMITH", "DWAYNE"), c("SMYTH", "DUANE"))

candidates_a <- c("John Smith", "Mary Jones")
candidates_b <- c("Jon Smith", "Mary-Anne Jones", "Someone Else")
jaro_winkler_matrix(candidates_a, candidates_b)

# Token reordering — two of three tokens match exactly, the third is close
jaro_winkler_tokens("Kyle John Haynes", "John Kylie Haynes")

# Punctuation/spacing variants of the same name
jaro_winkler_tokens(c("OBrien", "OBrien"), c("O Brien", "O'Brien"))
```

The full mechanics of how `jaro_winkler_tokens()` resolves reordered
tokens — and the `extra_penalty` argument for handling a stray extra
token — are in @sec-theory and the function reference (@sec-reference).

## Edit distance

`levenshtein()`, `osa_distance()`, `damerau_levenshtein()`, and `hamming()`
cover plain edits, restricted adjacent transpositions, unrestricted
Damerau-Levenshtein, and position-wise differences. Existing APIs compare
bytes by default; set `use_bytes = FALSE` for UTF-8 code points.

```{r}
#| label: editdist
levenshtein("kitten", "sitting")
osa_distance("ca", "abc")             # 3 under the OSA restriction
damerau_levenshtein("ca", "abc")      # 2 under unrestricted DL
levenshtein_similarity("kitten", "sitting")
levenshtein_within("kitten", "sitting", max_distance = 3)
hamming("karolin", "kathrin")
hamming("abc", "ab")             # Inf for unequal length, matching stringdist
```

## Fuzzy lookup without a full matrix

`fuzzy_match()` and `fuzzy_top_n()` scan a candidate table while retaining
only the requested matches. They default to Unicode code-point comparison
and avoid the potentially enormous all-pairs matrix.

```{r}
#| label: fuzzy-lookup
choices <- c("SMYTH", "JONAS", "JONES", "MITTEN", "SITTING")
fuzzy_match(c("SMITH", "JONES"), choices)
fuzzy_top_n("kitten", choices, method = "levenshtein", top_n = 2)
```

## Q-gram set similarity

`jaccard_index()`, `dice_coefficient()`, and `tversky_index()` compare two
strings' sets of overlapping length-`q` substrings ("q-grams") rather than
their character-by-character alignment — symmetric in token order and
insensitive to *where* a difference occurs, which makes them a useful
second signal alongside Jaro-Winkler/edit-distance.

```{r}
#| label: qgram
jaccard_index("night", "nacht")
dice_coefficient("night", "nacht")     # weights shared q-grams more heavily
tversky_index("night", "nacht", alpha = 1, beta = 1)  # == jaccard_index()
```

`tversky_index()` generalises both: `alpha = beta = 1` is Jaccard,
`alpha = beta = 0.5` is Dice, and asymmetric weights are useful when one
side is a query and the other a reference and the two kinds of mismatch
shouldn't be penalised equally.

## fuzzywuzzy-style ratios

`fuzz_ratio()`, `fuzz_partial_ratio()`, `fuzz_token_sort_ratio()`, and
`fuzz_token_set_ratio()` are vectorised ports of Python's
[fuzzywuzzy](https://github.com/seatgeek/fuzzywuzzy) package — useful if
you're porting a matching pipeline from Python, or just want its specific
flavour of fuzzy scoring (Ratcliff/Obershelp matching blocks, *not* edit
distance). All four return `0-100` like the originals.

```{r}
#| label: fuzz
fuzz_ratio("this is a test", "this is a test!")
fuzz_partial_ratio("fuzzy wuzzy was a bear", "wuzzy fuzzy was a bear")
fuzz_token_sort_ratio("fuzzy was a bear", "bear was a fuzzy")
fuzz_token_set_ratio("fuzzy was a bear", "fuzzy fuzzy bear was a bear")
```

## Putting it together

A typical pipeline normalises with `fchartr()`/`ftrimws()`, builds a
blocking key with `soundex()`/`nysiis()`/`format_date()`, then scores
candidates within each block with `jaro_winkler_matrix()` or
`jaro_winkler_tokens()`. @sec-workflow walks through exactly that, end to
end, on a small linkage example.
