# Algorithms under the hood {#sec-theory}

```{r}
#| label: setup
#| include: false
source("_setup.R")
has_recordlinkage <- requireNamespace("RecordLinkage", quietly = TRUE)
```

This chapter explains *why* the package is built the way it is, not just
*what* each function does (that's @sec-reference). Skip it if you just want
to call functions; read it if a benchmark number surprises you, or if
you're deciding whether to trust the package's behaviour on a pattern or
input shape you care about.

## Regex, fixed literals, and a safety net

R's default regex engine is TRE, which doesn't support lookaround,
possessive quantifiers, or atomic groups. `perl = TRUE` switches to PCRE,
which does — at the cost of being a backtracking engine, which means
certain patterns (nested quantifiers over ambiguous alternations, mainly)
can exhibit catastrophic exponential-time backtracking on adversarial
input.

`fast.string` uses **PCRE2** for regex (full modern syntax, via Rcpp, not
`base::grepl(perl = TRUE)`) and a prepared, byte-oriented searcher for the
`fixed = TRUE` literal-matching path. The fixed pattern is prepared once
per call: for searching, an empty pattern matches every non-`NA` string;
a one-byte pattern uses `memchr`; two- and three-byte patterns use `memchr`
followed by `memcmp`; and longer patterns use Boyer–Moore–Horspool.
Case-insensitive fixed matching folds bytes through a precomputed
256-entry table as they are compared, so it doesn't allocate a lowercased
copy of every input. The pattern is never interpreted as regex syntax on
this path, so there is no regex backtracking risk.

The catch with PCRE2 is that it remains a backtracking engine for genuine
regex patterns. What `fast.string` adds is a **syntax safety net**, not a
backtracking guarantee: every regex-accepting function scans the pattern
for PCRE-only syntax before taking the fast regex path —

```{r}
#| label: pcre-detect
#| eval: false
.has_pcre_only_syntax <- function(pattern) {
    base::grepl(
        paste0(
            "\\(\\?[=!]",       # (?= lookahead   (?! negative lookahead
            "|\\(\\?<[=!]",     # (?<= lookbehind  (?<! negative lookbehind
            "|\\(\\?>",         # (?> atomic group
            "|[*+?]\\+",        # *+  ++  ?+  possessive quantifiers
            "|\\(\\?P[=<]",     # (?P=  (?P<  named backref / group
            "|\\\\k[<']",       # \k<name>  \k'name'  named backreference
            "|\\(\\?R\\)",      # (?R) full-pattern recursion
            "|\\(\\?[0-9]",     # (?1) (?2) … numbered group recursion
            "|\\(\\?&"          # (?&name) named group recursion
        ),
        pattern, perl = TRUE
    )
}
```

— and if it finds any, and the caller didn't explicitly pass
`perl = TRUE`, the call transparently delegates to `base::grepl()`/
`base::gsub()` with `perl = TRUE` instead of running it through
`fast.string`'s own PCRE2 path, with a `cli::cli_inform()` notice
explaining why:

```{r}
#| label: pcre-fallback
fgrepl("(?<=foo)bar", c("foobar", "bazbar"), perl = FALSE)
```

This exists because `fast.string`'s PCRE2 binding and base R's PCRE
binding are two different code paths with two different sets of supported
constructs at the edges — the safety net means a pattern either runs
correctly on the fast path, or falls back to a path R has already battle-
tested for that exact syntax. It never silently mismatches.

`fcount()` inherits the same net, plus one of its own. Counting means
looping the match, so it also has to define what happens at the boundaries:
matches are counted **non-overlapping** (`fcount("aa", "aaaa")` is `2`), and
an empty pattern — whose base R semantics are counted in characters, while
the fixed engine works in bytes — is delegated to `base::gregexpr()`
outright rather than answered in bytes for UTF-8 input.

## Parallelism: RcppParallel, not `parallel`/`future`

`parallel::mclapply()` and `future::plan(multisession)` parallelise across
**R processes**: each worker is a separate R session, so every call pays
process-management and (for anything that isn't fork-based) serialisation
overhead — typically milliseconds at minimum, often more. That's
negligible if each task takes seconds; it dominates if each task is "trim
this one string."

`fast.string` instead uses [RcppParallel](https://rcppcore.github.io/RcppParallel/)'s
`parallelFor`, which partitions a single C++ loop across an Intel TBB
thread pool **inside one R call**, with no process boundary and no
serialisation step — the per-call dispatch overhead is microseconds, not
milliseconds. The trade-off is that this parallelism lives entirely inside
the C++ implementation: it's not something you can apply to arbitrary R
code, only to the specific loops this package has written in C++.

Because even microsecond dispatch overhead isn't free, every parallel
function applies a size threshold before bothering: typically `n >= 1000`
elements for pairwise vector operations, `n * m >= 10000` cells for matrix
operations (`jaro_winkler_matrix()`). Below the threshold, the function
just runs the loop on the calling thread — measurably faster than paying
thread-pool dispatch cost for, say, ten comparisons. @sec-benchmarks shows
this crossover explicitly with small-`n` timings.

Functions that expose `nthreads` treat it as a **per-call cap**, without
changing RcppParallel's process-wide options. `NULL` leaves scheduling to
the RcppParallel default, an explicit value limits that call to at most
that many threads, and `nthreads = 1` runs strictly serially. Dispatch also
limits concurrency to the useful work available, so asking for 16 threads
doesn't create 16-way overhead for a range with only a few useful chunks.
String-utility, date, and phonetic functions preserve their public
signatures and use the default scheduler for their own calls.

## Jaro and Jaro-Winkler similarity

Jaro similarity between strings $s_1$ (length $l_1$) and $s_2$ (length
$l_2$) is:

$$
\text{Jaro}(s_1, s_2) = \frac{1}{3}\left(\frac{m}{l_1} + \frac{m}{l_2} + \frac{m - t}{m}\right)
$$

where $m$ is the number of *matching* characters — characters common to
both strings within a sliding window of $\lfloor \max(l_1, l_2)/2 \rfloor - 1$
positions — and $t$ is the number of *transpositions* (matched characters
that appear in a different relative order) divided by two. Jaro-Winkler
adds a bonus for a shared prefix (up to 4 characters), scaled by `p`
(default `0.1`, the standard value):

$$
\text{JW}(s_1, s_2) = \text{Jaro}(s_1, s_2) + \ell \cdot p \cdot (1 - \text{Jaro}(s_1, s_2))
$$

where $\ell$ is the length of the common prefix, capped at 4. This is why
`jaro_winkler("SMITH", "SMYTH", p = 0.1)` scores higher than the same pair
with `p = 0` — `SM` is a shared prefix that the Winkler boost rewards, on
the reasoning that people are more likely to make typos later in a word
than at the start.

### The bit-parallel trick

The textbook Jaro algorithm, for each character in $s_1$, linearly scans
the matching window in $s_2$ for the first unclaimed equal character — an
$O(l_1 \times \text{window})$ double loop. `fast.string` instead maintains,
per worker thread, a 256-entry table of 64-bit bitmasks: `char_mask[c]` has
bit $j$ set if `s2[j] == c`. Finding the first unclaimed match for
`s1[i]` then becomes:

```cpp
uint64_t cand = char_mask[(unsigned char)s1[i]] & window_mask & ~used_mask;
if (cand) {
    int j = ctz64(cand);  // lowest set bit = leftmost match
    ...
}
```

— a handful of bitwise operations and one count-trailing-zeros instruction,
replacing the inner linear scan entirely. This turns the
$O(l_1 \times \text{window})$ comparison into $O(l_1 + l_2)$, and applies
to any pair of strings up to 64 bytes each (covering the overwhelming
majority of names and addresses). Longer strings fall back to a
straightforward `bool[]`-array scalar implementation — still correct, just
without the bitmask speedup.

The `char_mask` table itself is `thread_local` and persists across calls
within a worker thread (only the handful of bytes actually touched per
call get cleared afterward), so the steady-state per-call setup cost is
$O(l_2)$, not $O(256)$.

### Token-aware comparison: `jaro_winkler_tokens()`

Plain Jaro-Winkler compares strings as one ordered sequence of characters,
so `"Kyle John Haynes"` vs. `"John Kylie Haynes"` scores badly — the
*characters* moved, even though the *words* are almost all present.
`jaro_winkler_tokens()` scores every pair two different ways and keeps
whichever is higher:

1. **Token alignment.** Split both strings on whitespace, score every
   token in `a` against every token in `b`, then greedily pair off tokens
   in descending order of similarity (repeatedly pick the best remaining
   unused pair). This is a cheap approximation of an optimal bipartite
   assignment — not guaranteed optimal in the general case, but for the 2-5
   tokens typical of a name or address, it agrees with the true optimum
   for all but pathological inputs. By default, the summed score divides
   by `max(n_tokens_a, n_tokens_b)`, so missing/extra tokens dilute the
   result; the `extra_penalty` argument switches to a scheme that instead
   averages over only the *matched* tokens and subtracts a fixed penalty
   per unmatched token — see @sec-reference for the worked example with
   a stray middle initial.
2. **Collapsed.** Both strings with whitespace removed, compared as one
   string. This catches cases where the *only* difference is where the
   token boundary falls — `"OBrien"` vs. `"O Brien"` — which token
   alignment alone would under-score (one token vs. two, diluting the
   average) even though they're clearly the same name.

```{r}
#| label: tokens-demo
jaro_winkler_tokens("Kyle John Haynes", "John Kylie Haynes")  # token alignment wins
jaro_winkler_tokens("OBrien", "O Brien")                      # collapsed framing wins
```

Both framings reuse the same underlying Jaro-Winkler primitive and the
same parallelised, per-row execution as `jaro_winkler()` itself — tokenising,
building the token similarity matrix, the greedy assignment, and the
collapsed comparison all happen in one C++ pass over the whole input
vector.

## Soundex and NYSIIS

Both are **phonetic encoding** schemes: they map a name to a short code
such that names that *sound* alike (even if spelled differently) tend to
map to the same code. They're used as **blocking keys** — group records by
phonetic code first, then only run the expensive Jaro-Winkler comparison
*within* a group, instead of against every other record in the dataset.

**Soundex** (American/Census Bureau rules, implemented here) keeps the
first letter verbatim, then maps the remaining consonants to digits
(`B,F,P,V → 1`; `C,G,J,K,Q,S,X,Z → 2`; `D,T → 3`; `L → 4`; `M,N → 5`;
`R → 6`), drops vowels, and pads/truncates to exactly 4 characters.
Adjacent letters mapping to the *same* digit collapse to one — and `H`/`W`
are **transparent** for that adjacency check, not just dropped. That
transparency rule is the entire reason `"Ashcraft"` encodes as `A261` and
not `A2261`: the `S` and `C` are separated by an `H`, but `H` doesn't break
their adjacency, so they collapse to a single `2`.

```{r}
#| label: soundex-demo
soundex(c("Ashcraft", "Ashcroft", "Robert", "Rupert"))
```

```{r}
#| label: soundex-cross-pkg
#| eval: !expr has_recordlinkage
cat("fast.string::soundex:  ", fast.string::soundex("Ashcraft"), "(treats H as transparent)\n")
cat("RecordLinkage::soundex:", RecordLinkage::soundex("Ashcraft"), "\n")
```

This H/W-transparency rule is exactly the kind of edge case where Soundex
*implementations* genuinely disagree — some treat `H`/`W` as transparent
(the canonical Census Bureau / Knuth *TAOCP* rule, and what `fast.string`
follows), others drop them like any other consonant-adjacent vowel and get
a different code for words like `"Ashcraft"`. Neither is "more correct"
in any normative sense — Soundex was never formally standardised down to
this level of detail — but it's worth knowing which rule a given package
follows before assuming two Soundex columns from different tools are
comparable.

**NYSIIS** is more aggressive: it applies leading/trailing letter
transforms first (e.g. a name starting with `MAC` becomes `MCC`, one ending
in a vowel + `Y` drops the `Y`), then a per-letter substitution table with
adjacent-duplicate collapsing, then truncates to 6 characters. It tends to
produce fewer false-positive collisions than Soundex on typical English
surnames, at the cost of a heavier per-element rule table — which is the
direct explanation for why `nysiis()` benchmarks slower than `soundex()`
in @sec-benchmarks despite both running through the same parallel
dispatch.

### Refined Soundex and Cologne

Both are answers to the same complaint about classic Soundex — a
four-character code is very coarse — from opposite directions.

**Refined Soundex** (the US-English mapping from Apache Commons Codec)
keeps the first letter verbatim, then emits a class digit for *every*
letter of the name — the first one included, and vowels included, which
classic Soundex drops outright. Adjacent identical digits still collapse to
one, but nothing is truncated or padded, so the code grows with the name.
The mapping is nine consonant classes rather than Soundex's six
(`B,P → 1`; `F,V → 2`; `C,K,S → 3`; `G,J → 4`; `Q,X,Z → 5`; `D,T → 6`;
`L → 7`; `M,N → 8`; `R → 9`), with vowels and `H`/`W`/`Y` sharing `0`.
`"Robert"` therefore encodes as `R901096`, not `R163`: two names must now
agree in phonetic class at every position to collide, not just in their
first three significant consonants.

**Cologne phonetic** (*Kölner Phonetik*, Postel 1969) is built for German
rather than English. Its distinguishing feature for linkage work is
explicit **umlaut folding** — `Ä`/`Ö`/`Ü` map to their base vowels and
`ß`/`ẞ` to `SS`, decoded from UTF-8 before encoding — so `"Müller"` and
`"Mueller"` reach the same code by rule rather than by accident. Every
other scheme in this package is ASCII-oriented and simply skips the umlaut
byte pair, which is why `refined_soundex()` gives `"Müller"` and
`"Mueller"` different codes.

The practical consequence is block size. Over a list of surnames chosen to
stress each scheme:

```{r}
#| label: phonetic-selectivity
surnames <- c("Smith", "Smyth", "Schmidt", "Schmitt", "Sneath", "Sandy",
              "Robert", "Rupert", "Rubert", "Ashcraft", "Ashcroft",
              "Lee", "Law", "Loewe", "Baker", "Barker", "Hansen", "Hanson",
              "Jackson", "Jaxon", "Tymczak", "Pfister", "Meier", "Meyer",
              "Mayer", "Mueller", "Müller", "Miller", "Moller", "Malloy")

vapply(
    list(soundex = soundex, refined_soundex = refined_soundex,
         cologne = cologne, nysiis = nysiis),
    function(f) length(unique(f(surnames))),
    integer(1)
)
```

More distinct codes over the same input means smaller blocks, fewer
candidate pairs, and more missed matches — the recall/precision dial, in
one number. Soundex puts all six of
`"Smith"`/`"Smyth"`/`"Schmidt"`/`"Schmitt"`/`"Sneath"`/`"Sandy"` in a
single block; Refined Soundex splits them three ways:

```{r}
#| label: phonetic-selectivity-detail
s <- c("Smith", "Smyth", "Schmidt", "Schmitt", "Sneath", "Sandy")
data.frame(name = s, soundex = soundex(s), refined = refined_soundex(s),
           cologne = cologne(s))
```

Neither setting is correct in the abstract. The strategy that usually
works is to run more than one and union the candidate blocks, which buys a
coarse scheme's recall while keeping a fine scheme's precision on the
records the coarse one over-merges — @sec-workflow does exactly that with
`refined_soundex()` and `cologne()` together.

Two cross-package notes, in the same spirit as the Soundex `H`/`W`
disagreement above. Both matter if you are comparing a `fast.string`
phonetic column against one produced by another tool.

- **`refined_soundex()` does not truncate.** `phonics::refinedSoundex()`
  caps the code at 10 characters by default (`maxCodeLen = 10L`), so long
  names produce a shorter, coarser key there. Raise `maxCodeLen` and the
  two agree exactly; leave it at the default and `phonics` will merge
  records this package keeps apart.
- **Cologne's specification leaves two rules open**, and implementations
  read them differently. `cologne()` agrees with `phonics::cologne()` on
  the canonical reference names (`"Müller-Lüdenscheidt"`, `"Wikipedia"`,
  `"Breschnew"`, the `"Meier"`/`"Mayr"`/`"Maier"`/`"Meyer"` group), but the
  two diverge on a few per cent of random uppercase strings — mainly where
  an `H` sits between two letters that code to the same digit. This package
  treats the skipped `H` as still separating them (`cologne("BHB")` is
  `"11"`, and this is pinned in `tests/testthat/test-phonetic-extended.R`),
  where `phonics` collapses them to `"1"`. `phonics::cologne()` also
  returns `NA` with a warning on umlauts rather than folding them.

As with Soundex, neither reading is normatively "correct" — the point is
to know which one your column was built with before joining on it.

## Double Metaphone and Caverphone

`soundex()`/`nysiis()` were designed around English/American surnames;
`double_metaphone()` and `caverphone()` are the package's answer for names
that fall outside that design target.

**Double Metaphone** (Lawrence Philips, 2000) returns *two* codes per
word — `primary` and `secondary` — because some spellings genuinely have
more than one plausible pronunciation: a `C` could be a hard `K` sound or
a soft `S` sound depending on context, a `G` could be hard or soft, and so
on. Rather than guessing one and risking a missed match, the algorithm
emits both, and two names are usually considered a phonetic match if *any*
of one side's codes equals *any* of the other's. This package's
implementation is ported from the structure of the widely-used Apache
Commons Codec Java implementation and cross-checked against its published
unit-test vectors (`tests/testthat/test-double_metaphone.R`) rather than
hand-derived, since the ruleset is large (dozens of context-sensitive
branches per letter) and easy to get subtly wrong from a written
description alone.

```{r}
#| label: dmetaphone-demo
double_metaphone(c("Catherine", "Kathryn", "Smith", "Schmidt"))
```

`"Catherine"`/`"Kathryn"` sharing a code despite no letters lining up
positionally is the kind of match Soundex/NYSIIS — which work
letter-by-letter from the start of the word — structurally can't catch.

One implementation detail worth calling out: unlike `soundex()`/`nysiis()`,
which strip the input down to letters only before encoding, Double
Metaphone keeps internal whitespace and punctuation (only the ends are
trimmed, and the text uppercased). The algorithm's own rules key off
literal multi-character fragments like `"VAN "`/`"SAN "`/`"SCH"`,
*including the space* — stripping it first would silently break those
rules.

**Caverphone 2.0** (Caversham Project, University of Otago, 2004) takes a
completely different shape: instead of a branching per-letter ruleset, it's
a long fixed *chain* of substring transforms (originally specified as a
sequence of regex replacements) — strip non-letters, handle a handful of
silent-letter prefixes, replace `c`/`q`/`x`/`v`/`d`/`b`/`z` with their
phonetic equivalents, collapse vowel runs to a placeholder digit, then
clean up trailing consonants — always producing exactly 10 characters
(padded with `1`s). It was designed for, and is still widely used on, New
Zealand electoral-roll names, but works as a general-purpose alternative
blocking key anywhere. This package's implementation is ported
step-for-step from Apache Commons Codec's `Caverphone2` (again
cross-checked against its published test vectors,
`tests/testthat/test-caverphone.R`) including a documented quirk in the
original 2004 specification — the `"enough"`/`"trough"` prefix rules are
applied in an order that looks like a copy-paste duplication in the
original paper — preserved here rather than "corrected", since the point
of implementing a named phonetic algorithm is to match what every other
implementation of *that algorithm* produces.

```{r}
#| label: caverphone-demo
caverphone(c("Peter", "Tedder", "Stevenson"))
```

Trying more than one phonetic key (Soundex *and* NYSIIS *and* Double
Metaphone *and* Caverphone, unioning the candidate blocks each produces)
is a reasonable strategy when recall matters more than a small increase in
block size — different rulesets group different near-miss spellings
together, so one catches matches another misses.

## Edit distance: Levenshtein, OSA, and Damerau-Levenshtein

`levenshtein()`, `osa_distance()`, and `damerau_levenshtein()` are classic distance
metrics — minimum number of single-character edits to turn one string into
another. The textbook algorithm is an $O(l_1 \times l_2)$ dynamic-programming
table; `stringdist` (and most other implementations) use exactly that.

`levenshtein()` instead uses **Myers' (1999) bit-vector algorithm**: the
same family of trick as this package's bit-parallel Jaro
(@sec-theory's earlier section) — pack the *shorter* string's character
positions into 64-bit masks (one mask per distinct byte value), then scan
the *longer* string against those masks using a handful of bitwise
operations per character instead of an inner loop over the DP row. That
turns the usual $O(l_1 \times l_2)$ table into $O(l_1)$ word operations
whenever the shorter string fits in 64 bits — true for the overwhelming
majority of names/addresses. Pairs where even the shorter string exceeds
64 bytes fall back to a conventional rolling-row DP.

`osa_distance()` additionally treats an adjacent-character
transposition (`"ab"` → `"ba"`) as a single edit rather than two
substitutions — useful for typos, which are disproportionately adjacent
swaps. There's no equivalently compact bit-parallel formulation for the
transposition case that this package implements, so it uses a
three-rolling-row DP (transposition needs to look back two rows, not one,
unlike plain Levenshtein's one-row recurrence). This matches
`stringdist::stringdist(method = "osa")`: no substring is edited more than
once. `damerau_levenshtein()` uses the unrestricted Lowrance-Wagner
recurrence, matching `stringdist(method = "dl")`; it tracks the last row in
which each symbol occurred so multiple edits may involve the same substring.

```{r}
#| label: editdist-demo
levenshtein("kitten", "sitting")
osa_distance("ca", "abc")         # 3
damerau_levenshtein("ca", "abc")  # 2
```

The compatibility default is byte comparison. With `use_bytes = FALSE`,
inputs are converted and validated as UTF-8 on the calling R thread, then
workers compare immutable code-point arrays. ASCII inputs still use the
original bit-parallel byte kernels. Code-point mode does not normalize
canonically equivalent spellings and does not combine grapheme clusters.

## Q-gram similarity: Jaccard, Dice, Tversky, cosine

A different family of similarity entirely: instead of an alignment between
two character sequences, split each string into its overlapping length-`q`
substrings ("q-grams" — bigrams for the default `q = 2`) and compare those.
This is insensitive to *where* a difference occurs, which makes it a useful
complementary signal to edit-distance-style metrics rather than a
replacement for them.

`jaccard_index()`, `dice_coefficient()`, and `tversky_index()` all reduce
to the same underlying triple — `|A|`, `|B|`, `|A ∩ B|` — combined three
different ways:

- Jaccard: $|A \cap B| / |A \cup B|$
- Dice: $2|A \cap B| / (|A| + |B|)$ — always $\geq$ Jaccard for the same pair
- Tversky: $|A \cap B| / (|A \cap B| + \alpha|A \setminus B| + \beta|B \setminus A|)$,
  which reduces to Jaccard at $\alpha = \beta = 1$ and Dice at
  $\alpha = \beta = 0.5$

The speed advantage over `stringdist::stringdist(method = "jaccard")`
comes from two things: for q-gram lengths up to 8 bytes (covering the
default `q = 2` and the overwhelming majority of practical use), each
q-gram packs exactly into a `uint64_t` — one shift-and-OR per byte — so the
q-gram set is built and deduplicated as plain sorted integers, with no
string allocation and no hash table. `stringdist` instead tabulates
q-grams generically (to support arbitrary `q`, including very long ones)
through R-level dispatch per pair.

```{r}
#| label: qgram-demo
jaccard_index("night", "nacht")
dice_coefficient("night", "nacht")
```

### Cosine: profiles, not sets

`cosine_similarity()` breaks the pattern above. The other three reduce each
string to a *set* — a q-gram that occurs four times counts exactly as much
as one that occurs once. Cosine keeps the counts, treats each string as a
vector in q-gram space, and returns the cosine of the angle between them:

$$\cos(A, B) = \frac{\sum_g A_g B_g}{\sqrt{\sum_g A_g^2}\sqrt{\sum_g B_g^2}}$$

where $A_g$ is the number of times q-gram $g$ occurs in the first string.
Implementation-wise this is the same sorted-packed-integer machinery, with
one change: the q-gram vector is sorted but *not* deduplicated, and the
merge pass counts run lengths on each side to accumulate
$\sum A_g^2$, $\sum B_g^2$, and the dot product in a single sweep. Where
the set metrics normalise by set sizes, cosine normalises by the product
of the two profile magnitudes.

```{r}
#| label: cosine-demo
a <- c("aaaa", "BANANA", "night")
b <- c("aaab", "BANANAS", "nacht")
data.frame(a, b,
           jaccard = jaccard_index(a, b),
           dice    = dice_coefficient(a, b),
           cosine  = cosine_similarity(a, b))
```

`"aaaa"` and `"aaab"` share exactly one of two distinct bigrams, so Jaccard
reports `0.5` and cannot say more. Cosine sees three occurrences of `"aa"`
against two and scores the pair far higher. Which behaviour you want is a
data question: for names, near-duplicate q-grams are rare and the metrics
mostly agree; for the numeric identifiers, codes, and padded fields that
sit alongside names in a linkage file, repetition is the norm and the set
metrics are throwing away most of the signal.

Because they are computed from different information, cosine and
Jaro-Winkler make a reasonable *pair* of features for a linkage decision —
@sec-workflow scores candidates on both and treats a wide disagreement
between them as something to look at rather than average away.

## fuzzywuzzy-style ratios: Ratcliff/Obershelp matching blocks

`fuzz_ratio()`, `fuzz_partial_ratio()`, `fuzz_token_sort_ratio()`, and
`fuzz_token_set_ratio()` port Python's
[fuzzywuzzy](https://github.com/seatgeek/fuzzywuzzy) package — useful when
migrating a matching pipeline from Python, or when you specifically want
fuzzywuzzy's flavour of scoring rather than edit distance or Jaro-Winkler.

The core of all four is the **Ratcliff/Obershelp algorithm** — the same
algorithm behind Python's `difflib.SequenceMatcher`, which is what
fuzzywuzzy itself falls back to whenever the optional `python-Levenshtein`
C extension isn't installed (this package implements that always-available
reference behaviour, not the optional Levenshtein-based shortcut). It
works by recursively finding the *longest common contiguous matching
block* between two strings, then recursing into the unmatched left and
right remainders on each side — repeating until no more matches are
found. The total length of every matched block, `M`, gives a similarity
ratio of `2*M / (len(a) + len(b))`. This is a meaningfully different
notion of "similar" from edit distance: it rewards long shared *runs*
rather than penalising every single-character difference equally, so a
short insertion in the middle of an otherwise-identical string costs far
less here than under Levenshtein.

```{r}
#| label: fuzz-demo
fuzz_ratio("this is a test", "this is a test!")
fuzz_partial_ratio("fuzzy wuzzy was a bear", "wuzzy fuzzy was a bear")
```

`fuzz_partial_ratio()` builds on the same matching-blocks primitive: align
the *shorter* string against every position the longer one offers (via
those same matching blocks) and keep the best full-ratio alignment —
useful when one string is plausibly a true substring of the other, padded
with unrelated text on either side. `fuzz_token_sort_ratio()` and
`fuzz_token_set_ratio()` layer tokenisation on top (sort-and-rejoin, or
set-intersection-and-leftovers respectively) before running the same
ratio underneath, the same way `jaro_winkler_tokens()` layers tokenisation
on top of plain Jaro-Winkler.

## Fast, fixed-format dates

`format_date()`/`fas.Date()` deliberately don't replicate
`base::format.Date()`/`base::as.Date()`'s full generality. Base R's date
functions go through `strptime()`/`strftime()`, which are locale-aware and
support arbitrary `%`-format strings — flexibility that costs real time
when you're formatting the same fixed shape (say, ISO `YYYY-MM-DD`) for
five million rows. `fast.string`'s date functions instead support exactly
four fixed formats (`iso`, `compact`, `dmy`, `ymd_slash`) implemented as
direct digit-to-character writes with zero-padding — no locale lookup, no
calendar-math beyond what's needed to validate month (1-12) and day
(1-31) ranges. There's deliberately *no* days-in-month or leap-year check,
so `fas.Date("2024-02-30", "iso")` parses without error — trading
strictness for speed, on the assumption that upstream data has already
been validated or that downstream code will catch the implausible date
some other way. If you need full calendar validation, `base::as.Date()` is
still the right tool; `fas.Date()` is for when you already trust the input
shape and just want it parsed fast.

### Timestamps: `fas.POSIXct()` and `format_datetime()`

The timestamp pair applies the same fixed-shape principle one level up.
`base::as.POSIXct()` carries the same locale-aware `strptime()` machinery
as `as.Date()`, plus timezone resolution on top. `fas.POSIXct()` works
entirely in UTC seconds since the epoch, converting the calendar fields
arithmetically, and supports exactly four shapes (`iso`, `rfc3339`,
`compact`, `iso_offset`).

Two design choices are worth knowing before relying on them.

**These functions validate more than the date functions, not less.** Where
`fas.Date()` checks only that month and day fall in range, `fas.POSIXct()`
validates the full calendar — days-in-month and leap years included — and
returns `NA` for anything impossible. The asymmetry is deliberate: a
timestamp column is usually a load/event stamp that a pipeline reasons
about (ordering, watermarks, retention windows), so a silently rolled-over
value is a worse failure there than in a date-of-birth field that a human
will eyeball.

```{r}
#| label: datetime-strictness
fas.Date("2023-02-29", "iso")        # rolls over
fas.POSIXct("2023-02-29 00:00:00")   # NA
```

**`"iso_offset"` is an offset, not a timezone.** On input, an offset is
subtracted to normalise the value to UTC; on output, `format_datetime()`
adds one fixed offset to every element. There is no daylight-saving rule
anywhere in this path, and no zone name is ever resolved. For stamping a
batch whose offset you know, that is exactly right and very cheap. For a
series that crosses a DST boundary, or for anything that must round-trip
through a named zone like `"Australia/Sydney"`, use `base::format()` and
`base::as.POSIXct()` — this is a case where the fast version isn't a
drop-in replacement, it's a different (narrower) contract.

```{r}
#| label: datetime-offset
fas.POSIXct("2024-06-18T09:15:00+10:00", "iso_offset")   # stored as UTC
format_datetime(fas.POSIXct("2024-06-18 09:15:00"),
                "iso_offset", offset = "+10:00")          # written back out
```
