# A worked record-linkage pipeline {#sec-workflow}

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

This chapter strings every piece from @sec-quickstart together into a
single, realistic linkage pipeline: an incoming batch of records needs to
be matched against a reference database. It's the same shape of problem
`examples_data_linkage.R` in the repo demonstrates, written out here as a
narrative.

The batch arrives the way batches actually arrive — names in mixed case
with punctuation, dates as strings in a local format, a load timestamp per
row, a junk record that should never have been sent, and the same person
submitted twice. Getting from that to a defensible set of links is eight
steps, and most of the functions in this package show up in one of them.

## The data

```{r}
#| label: data
incoming <- data.frame(
    id       = 1:7,
    name     = c("  JOHN O'BRIEN  ", "MARY-JANE SMITH", "ROBERT BROWN",
                 "Kyle John Haynes", "Hans Müller", "UNKNOWN 00000000",
                 "JOHN OBRIEN"),
    dob      = c("15/03/1985", "22/07/1990", "01/12/1975", "05/01/1988",
                 "09/09/1971", "01/01/1900", "15/03/1985"),
    received = c("2024-06-18 09:15:00", "2024-06-18 09:17:42",
                 "2024-06-18 11:02:10", "2024-06-18 11:45:00",
                 "2024-06-18 12:00:00", "2024-06-18 12:30:00",
                 "2024-06-18 14:20:00"),
    stringsAsFactors = FALSE
)

reference <- data.frame(
    ref_id = 101:106,
    name   = c("JON OBRIEN", "MARYANN SMITH", "ROB BROWN", "JOHN BRIEN",
               "Haynes John Kyle", "Hans Mueller"),
    dob    = as.Date(c("1985-03-14", "1990-07-20", "1975-12-02",
                       "1985-03-15", "1988-01-05", "1971-09-09")),
    stringsAsFactors = FALSE
)

incoming
reference
```

Seven incoming records, six reference records, deliberately messy:
inconsistent casing, stray whitespace, punctuation, typos, a near-duplicate
in the reference file (`"JOHN BRIEN"` vs. `"JON OBRIEN"`), a fully
reordered multi-token name (`"Kyle John Haynes"` vs. `"Haynes John Kyle"`),
a German name spelled two ways (`"Hans Müller"` vs. `"Hans Mueller"`), a
placeholder record (`"UNKNOWN 00000000"`), and one person submitted twice
in the same batch (ids 1 and 7).

## Step 1: profile before you match

The cheapest way to improve a linkage run is to not feed it rubbish.
`fcount()` counts non-overlapping matches per element, which makes
column-level quality rules one vectorised call each: how many digits are in
a name field (a name with digits is almost always a placeholder or a
concatenated identifier), and how many tokens a name has.

```{r}
#| label: profile
incoming$n_digits <- fcount("[0-9]", incoming$name)
incoming$n_tokens <- fcount(" +", ftrimws(incoming$name)) + 1L

incoming[, c("id", "name", "n_digits", "n_tokens")]
```

Record 6 is the placeholder. `fcount()` is the counting counterpart to
`fgrepl()` — same PCRE2 and prepared-fixed-string engines, same parallel
dispatch — and replaces the `vapply(gregexpr(...), length, ...)` idiom,
which allocates a match-position vector per row just to take its length and
needs two explicit branches to reproduce `fcount()`'s `NA`-and-zero
handling:

```{r}
#| label: profile-equivalence
identical(
    fcount("[0-9]", incoming$name),
    vapply(gregexpr("[0-9]", incoming$name),
           function(m) if (is.na(m[1])) NA_integer_
                       else if (m[1] < 0L) 0L else length(m),
           integer(1))
)
```

## Step 2: parse dates and timestamps

The batch carries dates as `DD/MM/YYYY` strings and a load timestamp per
row. `fas.Date()` and `fas.POSIXct()` parse one fixed, locale-free shape
each — no format guessing, no timezone-database lookup.

```{r}
#| label: parse
incoming$dob_date    <- fas.Date(incoming$dob, "dmy")
incoming$received_ts <- fas.POSIXct(incoming$received)

incoming[, c("id", "dob", "dob_date", "received", "received_ts")]
```

The two functions differ in strictness, and it matters when you're deciding
which one is allowed to silently accept bad input. `fas.Date()` validates
field ranges only (month 1-12, day 1-31), so an impossible calendar date
rolls over; `fas.POSIXct()` validates the full calendar, including month
lengths and leap years, and returns `NA`:

```{r}
#| label: parse-strictness
fas.Date("29/02/2023", "dmy")          # rolls over -- 2023 is not a leap year
fas.POSIXct("2023-02-29 08:00:00")     # NA
```

So `fas.POSIXct()` doubles as a validity check on the timestamp column, and
an `NA` there is a legitimate reason to quarantine a row.

## Step 3: normalise and quarantine

Strip punctuation and whitespace inconsistencies *before* comparing, so the
comparison step measures genuine name similarity rather than formatting
noise — then set aside the rows the profile step flagged.

```{r}
#| label: normalise
incoming$name_clean <- fgsub(" +", " ", fchartr("'-", "  ", ftrimws(incoming$name)))
reference$name_clean <- fchartr("'-", "  ", ftrimws(reference$name))

usable <- incoming$n_digits == 0L &
    !is.na(incoming$dob_date) &
    !is.na(incoming$received_ts)

quarantine <- incoming[!usable, ]
batch      <- incoming[usable, ]

batch[, c("id", "name_clean", "dob_date", "received_ts")]
nrow(quarantine)
```

(`jaro_winkler_tokens()` does its own internal punctuation stripping and
case-folding — see @sec-reference — but normalising up front keeps this
example explicit, and matters for `jaro_winkler()`/`jaro_winkler_matrix()`
and the q-gram metrics, which don't.)

## Step 4: collapse duplicates inside the batch

Ids 1 and 7 are the same person, sent twice. Deduplicating *before*
matching keeps a single incoming record from generating two competing links
to the same reference row. The parsed timestamp decides which copy
survives — last write wins.

```{r}
#| label: dedupe
batch$dedupe_key <- paste(fgsub(" ", "", toupper(batch$name_clean)), batch$dob_date)
batch[, c("id", "dedupe_key", "received_ts")]

batch <- batch[order(batch$dedupe_key, -as.numeric(batch$received_ts)), ]
batch <- batch[!duplicated(batch$dedupe_key), ]
batch <- batch[order(batch$id), ]

batch[, c("id", "name_clean", "dob_date", "received_ts")]
```

Record 1 (09:15) is dropped in favour of record 7 (14:20). Note that this
step only catches duplicates that normalise to the *same* string —
genuinely fuzzy within-batch duplicates need the same scoring machinery as
the linkage itself, run with `batch` on both sides.

## Step 5: block on more than one phonetic key

Comparing every incoming record against every reference record is
$O(n \times m)$ — fine at this toy scale, prohibitive at millions of rows.
**Blocking** groups records by a cheap, coarse key first and only compares
records *within* the same block.

The classic key is `soundex()` on a surname, but a single key inherits that
scheme's blind spots. Two of the additions here exist precisely to cover
different ones:

```{r}
#| label: phonetic-comparison
german <- c("Müller", "Mueller", "Miller", "Meier", "Meyer")
data.frame(
    name    = german,
    soundex = soundex(german),
    refined = refined_soundex(german),
    cologne = cologne(german)
)
```

Read that table in the direction of *how much each key merges*. Classic
`soundex()` does catch `"Müller"`/`"Mueller"` here — but only as a side
effect of discarding vowels entirely, which is also why it drops
`"Miller"` in with them and collapses `"Meier"`/`"Meyer"` to a single
`M600`. That bluntness is the thing you're trying to get away from at
scale: coarse keys make big blocks, and big blocks are what blocking
exists to avoid.

`refined_soundex()` is the precise end. It encodes vowels rather than
discarding them and doesn't truncate, so blocks get smaller — but the
umlaut isn't an ASCII vowel it can encode, so it splits `"Müller"`
(`M8709`) from `"Mueller"` (`M80709`) and on its own would never make that
pair a candidate. `cologne()` folds umlauts by design and gives both
`657`, while staying coarse enough to pull `"Miller"` in too.

Neither key is right on its own, and that's the point: running both and
unioning the candidate sets recovers the specific pair the precise key
misses without adopting the coarse key's block sizes everywhere else.
@sec-theory quantifies the selectivity difference across a wider surname
list.

Because names arrive in inconsistent token order, block on the phonetic
code of *every* token rather than a positional "surname" — a record is a
candidate if it shares any token code, within the same birth year.

```{r}
#| label: blocking
block_keys <- function(ids, names_vec, years, scheme) {
    toks  <- strsplit(names_vec, " +")
    codes <- scheme(unlist(toks))
    out <- data.frame(
        id    = rep(ids, lengths(toks)),
        key   = paste0(codes, ":", rep(years, lengths(toks))),
        token = unlist(toks),
        stringsAsFactors = FALSE
    )
    out[!is.na(codes) & nzchar(codes), ]
}

inc_year <- as.integer(format(batch$dob_date, "%Y"))
ref_year <- as.integer(format(reference$dob, "%Y"))

inc_keys <- rbind(
    transform(block_keys(batch$id, batch$name_clean, inc_year, refined_soundex),
              scheme = "refined_soundex"),
    transform(block_keys(batch$id, batch$name_clean, inc_year, cologne),
              scheme = "cologne")
)
ref_keys <- rbind(
    transform(block_keys(reference$ref_id, reference$name_clean, ref_year, refined_soundex),
              scheme = "refined_soundex"),
    transform(block_keys(reference$ref_id, reference$name_clean, ref_year, cologne),
              scheme = "cologne")
)
names(ref_keys)[c(1, 3)] <- c("ref_id", "ref_token")

head(inc_keys, 8)
```

The whole exploded token vector is encoded in one vectorised call per
scheme, so the per-token cost is a parallel C++ pass rather than an R-level
loop over records.

```{r}
#| label: candidates
hits  <- merge(inc_keys, ref_keys, by = c("key", "scheme"))
pairs <- unique(hits[, c("id", "ref_id")])
pairs <- pairs[order(pairs$id, pairs$ref_id), ]

nrow(pairs)
nrow(batch) * nrow(reference)   # the full cross product this replaces
pairs
```

Six candidate pairs instead of thirty. At this scale that saving is
cosmetic; the point is the ratio, which is what survives to five million
rows — and that every true match is still in the set.

Keeping the token alongside the key makes the blocking step auditable —
you can ask *why* any given pair is a candidate, which matters when you're
tuning keys and need to know which one is carrying the recall:

```{r}
#| label: provenance
unique(hits[hits$id == 5 & hits$ref_id == 106,
            c("scheme", "token", "ref_token", "key")])
```

Note what this says, and what it doesn't. The `"Hans Müller"` pair is
reachable under *both* schemes, because the given name is identical and
`refined_soundex()` blocks it happily. Cologne's contribution is the second
row: it is the only one of the two that also links the *surnames*
(`"Müller"`/`"Mueller"` → `657`). On this batch that's redundant — but
given names disagree across systems constantly (`"Hans"`/`"Johann"`/`"H"`),
and when they do, that surname key is the only thing keeping the pair
alive.

## Step 6: score the candidates

Score name similarity only among surviving candidates, with two
deliberately different metrics. `jaro_winkler_tokens()` compares names
token by token and is insensitive to token order by construction.
`cosine_similarity()` compares q-gram *frequency profiles* across the whole
string, so it is order-sensitive and weights repeated q-grams — an
independent signal rather than a restatement of the first.

```{r}
#| label: scoring
pairs$name_a  <- batch$name_clean[match(pairs$id, batch$id)]
pairs$name_b  <- reference$name_clean[match(pairs$ref_id, reference$ref_id)]
pairs$dob_gap <- abs(as.integer(
    batch$dob_date[match(pairs$id, batch$id)] -
    reference$dob[match(pairs$ref_id, reference$ref_id)]
))

pairs$jw  <- jaro_winkler_tokens(pairs$name_a, pairs$name_b, ignore_case = TRUE)
pairs$cos <- cosine_similarity(toupper(pairs$name_a), toupper(pairs$name_b), q = 2)

pairs[order(pairs$id, -pairs$jw),
      c("id", "name_a", "ref_id", "name_b", "jw", "cos", "dob_gap")]
```

The two scores disagree in informative ways. `"Kyle John Haynes"` vs.
`"Haynes John Kyle"` is a perfect `1.000` under token comparison — every
token has an exact partner — but `0.867` under cosine, which sees the
reordering as changed bigrams across token boundaries. That gap *is* the
information: a high `jw` with a lower `cos` says "same tokens, different
arrangement", which is a different kind of evidence from "same arrangement,
one typo".

## Step 7: decide, and admit ambiguity

Apply a cutoff to decide which candidate pairs are accepted as links. 0.85
is a common starting point for Jaro-Winkler on names; the right value
depends on your data and how costly false positives/negatives are relative
to each other — tune it against a labelled sample if you have one. The
date gap is a second field-level filter, tolerating a small typo in the day
or month without accepting an unrelated birth date.

```{r}
#| label: threshold
accepted <- pairs[pairs$jw >= 0.85 & pairs$dob_gap <= 2, ]
accepted <- accepted[order(accepted$id, -accepted$jw), ]
accepted$ambiguous <- table(accepted$id)[as.character(accepted$id)] > 1L

accepted[, c("id", "name_a", "ref_id", "name_b", "jw", "cos", "dob_gap", "ambiguous")]
```

Every incoming record links to its obvious match, including the fully
reordered `"Kyle John Haynes"` → `"Haynes John Kyle"` pair — the one case
in this dataset where plain `jaro_winkler()` would have scored well below
threshold and `jaro_winkler_tokens()` is doing real work — and the
`"Hans Müller"` → `"Hans Mueller"` pair, which clears the threshold at
`0.914` despite the umlaut, because token comparison is scoring the
characters rather than a phonetic key.

Record 7 is the honest case. It has two surviving candidates, `0.980` and
`0.973`, and *identical* cosine scores of `0.843`:

```{r}
#| label: margin
best   <- accepted[!duplicated(accepted$id), ]
second <- accepted[duplicated(accepted$id), ]
best$margin <- best$jw - second$jw[match(best$id, second$id)]

best[, c("id", "name_a", "ref_id", "name_b", "jw", "cos", "margin")]
```

A margin of 0.007 between the top two candidates is not a decision, it's a
coin toss with extra steps. The second metric doesn't break the tie here —
it agrees — and that agreement is itself the finding: `"JOHN OBRIEN"` is
genuinely close to both `"JOHN BRIEN"` and `"JON OBRIEN"`, and no threshold
on this evidence can tell you which. The right output is a review queue,
not a link. Routing thin-margin pairs to clerical review, rather than
letting `!duplicated()` silently pick the first row, is the difference
between a linkage you can defend and one you can't.

## Step 8: stamp the output

Downstream loads need to know which batch they're looking at.
`format_datetime()` writes the batch watermark without a timezone-database
lookup, in UTC or at one fixed offset for a local-time report:

```{r}
#| label: watermark
watermark <- max(batch$received_ts)

format_datetime(watermark, "rfc3339")                        # UTC, for the audit log
format_datetime(watermark, "iso_offset", offset = "+10:00")  # local time, for humans
format_datetime(watermark, "compact")                        # sortable partition key
```

`"iso_offset"` applies one fixed numeric offset to every value — it is not
a timezone. It has no daylight-saving rules, so a batch spanning a DST
boundary needs `base::format()` with a real timezone; for stamping a single
load with a known offset, this is the cheap and correct tool.

## Why the fast versions matter here

Steps 1 and 2 are the ones that quietly dominate a real batch load: they
touch every row of every column, before any of the interesting matching
work starts. A single timed run on this machine, over 300,000 rows:

```{r}
#| label: timing
set.seed(1)
n  <- 3e5
nm <- paste0(sample(c("John Smith", "Mary O'Brien", "Kyle J Haynes 12"), n, TRUE),
             " ", sample(999, n, TRUE))
ts <- format(as.POSIXct("2024-06-18 09:15:00", tz = "UTC") + sample(1e7, n), "%Y-%m-%d %H:%M:%S")

system.time(vapply(gregexpr("[0-9]", nm),
                   function(m) if (m[1] < 0L) 0L else length(m), integer(1)))
system.time(fcount("[0-9]", nm))

system.time(as.POSIXct(ts, tz = "UTC"))
system.time(fas.POSIXct(ts))
```

Both are single runs on one machine and will vary with cores and input
shape — @sec-benchmarks is the chapter with the reproducible harness, and
it scales the comparison step of this exact pipeline shape up to 3 million
rows.
