# Function reference {#sec-reference}

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

Every exported function, grouped the same way as the README. Each entry
gives the signature, every argument, the return value, and at least one
runnable example. This chapter is the detailed reference; @sec-quickstart
is the faster tour.

## Matching & substitution

### `fgrepl()`

Equivalent to `base::grepl()`, using PCRE2 and Intel TBB. Typically 5-20x
faster on large character vectors. See @sec-theory for the automatic
PCRE-only-syntax fallback to `base::grepl(perl = TRUE)`.

```r
fgrepl(pattern, x, ignore.case = FALSE, perl = FALSE,
       fixed = FALSE, useBytes = FALSE, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `pattern` | Character scalar. Pattern to search for. |
| `x` | Character vector. `NA` elements return `NA`. |
| `ignore.case` | Logical. Case-insensitive matching. |
| `perl` | Logical. If `TRUE`, skip the PCRE-only syntax check and run the parallel PCRE2 engine directly. |
| `fixed` | Logical. Treat `pattern` as a literal string (fastest path — uses RE2). |
| `useBytes` | Logical. Ignored; kept for signature compatibility with `base::grepl()`. |
| `nthreads` | Integer or `NULL`. Thread count; `NULL` uses all cores. |

**Returns:** logical vector the same length as `x`.

```{r}
fgrepl("^inv", c("invoice", "Invoice", "credit"), ignore.case = TRUE)
fgrepl("a.b", c("axb", "ab"), fixed = TRUE)  # literal "a.b" — no match for "axb"
```

### `fgrep()`

Equivalent to `base::grep()`: wraps `fgrepl()` and returns indices or
values.

```r
fgrep(pattern, x, ignore.case = FALSE, perl = FALSE, value = FALSE,
      fixed = FALSE, useBytes = FALSE, invert = FALSE, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `pattern`, `x`, `ignore.case`, `perl`, `fixed`, `useBytes`, `nthreads` | Same as `fgrepl()`. |
| `value` | Logical. Return matching elements instead of indices. |
| `invert` | Logical. Return non-matching indices/values instead. |

**Returns:** integer vector of indices, or a character vector when
`value = TRUE`.

```{r}
fgrep("^inv", c("invoice", "credit", "invoice2"))
fgrep("^inv", c("invoice", "credit", "invoice2"), value = TRUE)
```

### `fsub()`

Equivalent to `base::sub()`: first-match substitution. Supports `\\1`-`\\9`
capture groups and `\\U`/`\\L`/`\\E` case conversion in `replacement`.

```r
fsub(pattern, replacement, x, ignore.case = FALSE, perl = FALSE,
     fixed = FALSE, useBytes = FALSE, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `pattern` | Character scalar. Pattern to search for. |
| `replacement` | Character scalar. Replacement string. |
| `x` | Character vector. `NA` elements return `NA`. |
| `ignore.case`, `perl`, `fixed`, `useBytes`, `nthreads` | Same as `fgrepl()`. |

**Returns:** character vector the same length as `x`.

```{r}
fsub("-([0-9]+)$", "_\\1", c("invoice-001", "invoice-002"))
```

### `fgsub()`

Equivalent to `base::gsub()`: global substitution. Same capture-group and
case-conversion support as `fsub()`.

```r
fgsub(pattern, replacement, x, ignore.case = FALSE, perl = FALSE,
      fixed = FALSE, useBytes = FALSE, nthreads = NULL)
```

Arguments are identical to `fsub()`. **Returns:** character vector the
same length as `x`.

```{r}
fgsub("[aeiou]", "_", c("hello world", "fast string"), ignore.case = TRUE)
```

### `gsub_all()`

No base R equivalent: applies several patterns in a single function call,
either sequentially (chained, like repeated `fgsub()` calls — later
patterns can match text introduced by earlier replacements) or as one
combined left-to-right scan (`sequential = FALSE` — the first pattern to
match at each position wins; later replacements never see earlier ones'
output).

```r
gsub_all(patterns, replacements, x, fixed = FALSE, ignore.case = FALSE,
          sequential = TRUE, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `patterns` | Character vector of patterns to search for. |
| `replacements` | Character scalar or vector the same length as `patterns`. |
| `x` | Character vector. `NA` elements return `NA`. |
| `fixed` | Logical. Treat each pattern as a literal string. |
| `ignore.case` | Logical. Case-insensitive matching. |
| `sequential` | Logical (default `TRUE`). `TRUE` = chained passes; `FALSE` = single combined scan. |
| `nthreads` | Integer or `NULL`. Thread count. |

**Returns:** character vector the same length as `x`.

```{r}
gsub_all(c("Mr\\.", "Mrs\\."), c("Mister", "Missus"), c("Mr. Smith", "Mrs. Jones"))
```

## String utilities

All four functions in this group preserve `names(x)` and `NA` handling
exactly like their base R equivalents.

### `ftrimws()`

```r
ftrimws(x, which = c("both", "left", "right"), whitespace = "[ \t\r\n]")
```

| Argument | Description |
|---|---|
| `x` | Character vector. `NA` elements return `NA`. |
| `which` | One of `"both"`, `"left"`, `"right"`. |
| `whitespace` | Regex of characters to strip. Only the default `"[ \t\r\n]"` uses the fast path; any other value delegates to `base::trimws()`. |

**Returns:** character vector the same length as `x`, with `names(x)`
preserved.

```{r}
ftrimws(c("  padded  ", "\ttabbed\n"))
ftrimws("  left only", which = "left")
```

### `fsubstr()`

```r
fsubstr(x, start, stop)
```

| Argument | Description |
|---|---|
| `x` | Character vector. `NA` elements return `NA`. |
| `start`, `stop` | Integer (or numeric) vectors of length 1 or `length(x)`, 1-indexed, clamped to each string's bounds, recycled to `length(x)`. `NA` in either produces `NA` for that element. |

**Returns:** character vector the same length as `x`.

```{r}
fsubstr("NSWREC123456789", 1, 3)
fsubstr(rep("ABCDEFGH", 3), c(1, 4, 7), c(3, 6, 8))  # vectorised start/stop
```

### `fnchar()`

```r
fnchar(x, type = "chars", allowNA = FALSE, keepNA = NA)
```

| Argument | Description |
|---|---|
| `x` | Vector, coerced to character via `as.character()` if needed. |
| `type` | One of `"bytes"`, `"chars"`, or `"width"`. `"width"` delegates to `base::nchar()` (display width isn't parallelised). |
| `allowNA` | Logical. Ignored; the fast path never raises an encoding error regardless. |
| `keepNA` | Logical or `NA` (default). `NA`/`TRUE`: `NA` elements of `x` return `NA`. `FALSE`: they return `2L` (length of the string `"NA"`), matching base R's legacy behaviour. |

**Returns:** integer vector the same length as `x`, with `names(x)`
preserved.

```{r}
fnchar(c("café", "abc", NA))
fnchar(c("café", "abc"), type = "bytes")  # "é" is 2 bytes in UTF-8
```

### `fchartr()`

```r
fchartr(old, new, x)
```

| Argument | Description |
|---|---|
| `old`, `new` | Single strings with the same number of characters. Each character of `old` maps to the corresponding character of `new`. |
| `x` | Character vector. `NA` elements return `NA`. |

**Returns:** character vector the same length as `x`. Falls back to
`base::chartr()` automatically when `old`/`new` contain multi-byte
characters (the fast path is a flat byte lookup table, valid only for
single-byte/ASCII characters).

```{r}
fchartr("'-", "  ", c("O'Brien", "Smith-Jones"))
fchartr("aeiou", "AEIOU", "fast string")
```

## Dates

### `format_date()`

```r
format_date(x, format = c("iso", "compact", "dmy", "ymd_slash"))
```

| Argument | Description |
|---|---|
| `x` | A `Date` object or numeric vector of days since 1970-01-01. |
| `format` | `"iso"` (`YYYY-MM-DD`), `"compact"` (`YYYYMMDD`), `"dmy"` (`DD/MM/YYYY`), or `"ymd_slash"` (`YYYY/MM/DD`). |

**Returns:** character vector the same length as `x`.

```{r}
format_date(as.Date(c("2024-06-18", NA)), "compact")
```

### `format_date_parts()`

Pure string-building (zero-padding plus punctuation, no calendar math, no
`Date` object) — the inverse of `date_parts()`.

```r
format_date_parts(year, month, day, format = c("iso", "compact", "dmy", "ymd_slash"))
```

| Argument | Description |
|---|---|
| `year`, `month`, `day` | Numeric vectors, recycled to a common length. `NA`, non-integer-coercible, or out-of-width values (year outside 0-9999, month/day outside 0-99) produce `NA` for that element. No calendar validation — `month = 13` formats as given if it fits the field width. |
| `format` | Same four options as `format_date()`. |

**Returns:** character vector recycled to the common length of `year`,
`month`, `day`.

```{r}
format_date_parts(2024, 6, 18, "iso")
```

### `date_parts()`

```r
date_parts(x)
```

| Argument | Description |
|---|---|
| `x` | A `Date` object or numeric vector of days since 1970-01-01. |

**Returns:** a `data.frame` with integer columns `year`, `month`, `day`.

```{r}
date_parts(as.Date(c("2024-06-18", "1999-12-31")))
```

### `fas.Date()`

Not a drop-in for `base::as.Date()` — see @sec-theory for the speed/
strictness trade-off (no days-in-month or leap-year validation).

```r
fas.Date(x, format = c("iso", "compact", "dmy", "ymd_slash"))
```

| Argument | Description |
|---|---|
| `x` | Character vector. `NA` elements, and elements that don't match `format` exactly (wrong length/separators/non-digits) or have an out-of-range month/day, become `NA`. |
| `format` | Same four options as `format_date()`; `x` must be in exactly this one format. |

**Returns:** a `Date` vector the same length as `x`.

```{r}
fas.Date(c("18/06/2024", "31/12/1999", "not-a-date"), "dmy")
```

## Phonetic blocking keys

Both functions have no base R equivalent and are intended as blocking
keys ahead of fuzzy name matching (@sec-theory, @sec-workflow).

### `soundex()`

```r
soundex(x)
```

| Argument | Description |
|---|---|
| `x` | Character vector (coerced via `as.character()` if not already character — e.g. a factor read from CSV). `NA` elements, and elements with no alphabetic characters, return `NA`. |

**Returns:** character vector the same length as `x`, with `names(x)`
preserved; each element either `NA` or exactly 4 characters.

```{r}
soundex(c("Robert", "Rupert", "Ashcraft"))
```

### `nysiis()`

```r
nysiis(x)
```

| Argument | Description |
|---|---|
| `x` | Character vector (coerced via `as.character()` if needed). `NA` elements, and elements with no alphabetic characters, return `NA`. |

**Returns:** character vector the same length as `x`, with `names(x)`
preserved; each element either `NA` or up to 6 characters.

```{r}
nysiis(c("Robert", "Rupert", "Ashcraft"))
```

## Fuzzy string similarity

### `jaro_winkler()`

Vectorised pairwise Jaro-Winkler similarity. See @sec-theory for the
algorithm and the bit-parallel implementation.

```r
jaro_winkler(a, b, p = 0.1, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `a`, `b` | Equal-length character vectors. |
| `p` | Prefix scaling factor (default `0.1`, the standard value). |
| `nthreads` | Integer or `NULL`. Thread count. |

**Returns:** numeric vector of similarities in `[0, 1]`.

```{r}
jaro_winkler(c("MARTHA", "DWAYNE"), c("MARHTA", "DUANE"))
jaro_winkler("SMITH", "SMYTH", p = 0)    # no prefix bonus
jaro_winkler("SMITH", "SMYTH", p = 0.1)  # standard prefix bonus
```

### `jaro_winkler_matrix()`

All-pairs Jaro-Winkler similarity — every element of `a` against every
element of `b`.

```r
jaro_winkler_matrix(a, b, p = 0.1, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `a` | Character vector of length n (rows). |
| `b` | Character vector of length m (columns). |
| `p`, `nthreads` | Same as `jaro_winkler()`. |

**Returns:** numeric matrix of dimensions n × m.

```{r}
jaro_winkler_matrix(c("JOHN SMITH", "MARY JONES"), c("JON SMYTH", "MARY JONES", "JOHN SMITH"))
```

### `jaro_winkler_tokens()`

Token-reorder- and punctuation-aware Jaro-Winkler similarity. Full
mechanics in @sec-theory.

```r
jaro_winkler_tokens(a, b, p = 0.1, ignore_case = FALSE, strip = "['’-]",
                     extra_penalty = NULL, nthreads = NULL)
```

| Argument | Description |
|---|---|
| `a`, `b` | Equal-length character vectors. |
| `p` | Prefix scaling factor (default `0.1`). |
| `ignore_case` | Logical (default `FALSE`). Uppercase both sides before comparing. |
| `strip` | Single regex string matching punctuation to remove before tokenising, or `NULL` to skip. Default `"['’-]"` strips straight/curly apostrophes and hyphens. |
| `extra_penalty` | `NULL` (default) or a non-negative numeric scalar. `NULL`: unmatched tokens dilute the average (divide by `max(n_tokens_a, n_tokens_b)`). A number: average over only matched tokens, then subtract `extra_penalty` per unmatched token (floored at 0) — `0` ignores extra tokens outright. |
| `nthreads` | Integer or `NULL`. Thread count. |

**Returns:** numeric vector of similarities in `[0, 1]`. `NA` if either
side of a pair is `NA`.

```{r}
# Reordered tokens, one near-miss — rescued relative to plain jaro_winkler()
jaro_winkler_tokens("Kyle John Haynes", "John Kylie Haynes")
jaro_winkler("Kyle John Haynes", "John Kylie Haynes")

# Punctuation/spacing variants of the same name
jaro_winkler_tokens(c("OBrien", "OBrien"), c("O Brien", "O'Brien"))

# A stray extra token, penalised by default...
jaro_winkler_tokens("Kylie John ZZ Haynes", "Haynes John Kyle")
# ...vs. ignored outright with extra_penalty = 0
jaro_winkler_tokens("Kylie John ZZ Haynes", "Haynes John Kyle", extra_penalty = 0)
# ...vs. lightly discounted
jaro_winkler_tokens("Kylie John ZZ Haynes", "Haynes John Kyle", extra_penalty = 0.1)
```
