# Benchmarks at scale {#sec-benchmarks}

```{r}
#| label: setup
#| include: false
source("_setup.R")
suppressPackageStartupMessages({
    library(stringi)
    library(microbenchmark)
})
has_stringdist     <- requireNamespace("stringdist", quietly = TRUE)
has_recordlinkage  <- requireNamespace("RecordLinkage", quietly = TRUE)
has_phonics        <- requireNamespace("phonics", quietly = TRUE)
```

Every other chapter uses small inputs so the page renders quickly. This
chapter doesn't: every benchmark below runs on **3 million strings**,
because this package is built for exactly that scale — record linkage and
ETL pipelines where a "small" file is already six figures of rows. If a
function is only 10% faster at 3M rows, that's a real, honestly-reported
result, not a number picked to flatter the package.

Comparison baselines, used wherever they have a real equivalent:

- **base R** — the function this package's API mirrors.
- **stringi**/**stringr** — the standard fast-string ecosystem package for
  general string operations (stringr wraps stringi; benchmarks below call
  stringi directly).
- **stringdist** — `method = "jw"` is the closest equivalent to
  `jaro_winkler()`; `method = "lv"`/`"osa"`/`"dl"`/`"hamming"`/`"jaccard"`
  are the closest equivalents to `levenshtein()`/`osa_distance()`/
  `damerau_levenshtein()`/`hamming()`/`jaccard_index()`.
- **RecordLinkage** — `soundex()` and `jarowinkler()` are the closest
  equivalents to this package's phonetic and fuzzy-matching functions.
- **phonics** — `metaphone()` and `caverphone()` are throughput comparison
  points for `double_metaphone()`/`caverphone()`; they implement different
  algorithm revisions, so outputs aren't expected to match (the comparison
  is "a phonetic-coding R function over this vector", not output parity).

```{r}
#| label: versions
cat("fast.string: ", as.character(utils::packageVersion("fast.string")), "\n")
cat("stringi:     ", as.character(utils::packageVersion("stringi")), "\n")
if (has_stringdist)    cat("stringdist:  ", as.character(utils::packageVersion("stringdist")), "\n")
if (has_recordlinkage) cat("RecordLinkage:", as.character(utils::packageVersion("RecordLinkage")), "\n")
if (has_phonics)       cat("phonics:     ", as.character(utils::packageVersion("phonics")), "\n")
cat("R:           ", R.version.string, "\n")
cat("cores:       ", parallel::detectCores(), "\n")
```

```{r}
#| label: test-data
set.seed(1)
n <- 3e6

x <- stri_rand_strings(n, sample(2:40, n, replace = TRUE), pattern = "[A-Za-z0-9 ]")
na_idx    <- sample(n, 3000)
empty_idx <- sample(setdiff(seq_len(n), na_idx), 3000)
x[na_idx]    <- NA_character_
x[empty_idx] <- ""

cat(sprintf("n = %s strings (%d NA, %d empty, lengths 2-40)\n",
            format(n, big.mark = ",", scientific = FALSE), length(na_idx), length(empty_idx)))
```

A results table accumulates through the chapter and prints in full at the
end (@tbl-summary), alongside every individual comparison inline.

```{r}
#| label: results-accumulator
results <- list()
record <- function(op, baseline, baseline_s, fast_s) {
    baseline_s <- unname(baseline_s)
    fast_s <- unname(fast_s)
    results[[length(results) + 1]] <<- data.frame(
        operation = op, baseline = baseline,
        baseline_s = baseline_s, fast_s = fast_s,
        speedup = baseline_s / fast_s
    )
}
```

## Matching & substitution

```{r}
#| label: bench-grepl
t_base <- system.time(base::grepl("[0-9]{2}", x))["elapsed"]
t_fast <- system.time(fast.string::fgrepl("[0-9]{2}", x))["elapsed"]
t_stri <- system.time(stri_detect_regex(x, "[0-9]{2}"))["elapsed"]
cat(sprintf("base::grepl       %.3fs\nstringi::stri_detect_regex %.3fs\nfast.string::fgrepl %.3fs\n", t_base, t_stri, t_fast))
record("fgrepl (regex)", "base::grepl", t_base, t_fast)
```

```{r}
#| label: bench-gsub
t_base <- system.time(base::gsub("[0-9]+", "#", x, perl = TRUE))["elapsed"]
t_fast <- system.time(fast.string::fgsub("[0-9]+", "#", x))["elapsed"]
t_stri <- system.time(stri_replace_all_regex(x, "[0-9]+", "#"))["elapsed"]
cat(sprintf("base::gsub        %.3fs\nstringi::stri_replace_all_regex %.3fs\nfast.string::fgsub %.3fs\n", t_base, t_stri, t_fast))
record("fgsub (regex)", "base::gsub", t_base, t_fast)
```

```{r}
#| label: bench-gsub-all
patterns <- c("a", "e", "i", "o", "u")
repls    <- c("4", "3", "1", "0", "*")
t_base <- system.time({
    y <- x
    for (i in seq_along(patterns)) y <- base::gsub(patterns[i], repls[i], y, fixed = TRUE)
})["elapsed"]
t_fast <- system.time(fast.string::gsub_all(patterns, repls, x, fixed = TRUE, sequential = TRUE))["elapsed"]
cat(sprintf("base (5x chained gsub)  %.3fs\nfast.string::gsub_all   %.3fs\n", t_base, t_fast))
record("gsub_all (5 patterns)", "base (chained gsub)", t_base, t_fast)
```

## String utilities

```{r}
#| label: bench-str-utils
x_ws <- ifelse(is.na(x), NA_character_, paste0("  ", x, "\t\n"))

t_base <- system.time(base::trimws(x_ws))["elapsed"]
t_fast <- system.time(fast.string::ftrimws(x_ws))["elapsed"]
t_stri <- system.time(stri_trim_both(x_ws))["elapsed"]
cat(sprintf("base::trimws      %.3fs\nstringi::stri_trim_both %.3fs\nfast.string::ftrimws %.3fs\n", t_base, t_stri, t_fast))
record("ftrimws", "base::trimws", t_base, t_fast)

t_base <- system.time(base::substr(x, 1, 10))["elapsed"]
t_fast <- system.time(fast.string::fsubstr(x, 1, 10))["elapsed"]
cat(sprintf("base::substr      %.3fs\nfast.string::fsubstr %.3fs\n", t_base, t_fast))
record("fsubstr", "base::substr", t_base, t_fast)

t_base <- system.time(base::nchar(x, "chars"))["elapsed"]
t_fast <- system.time(fast.string::fnchar(x, "chars"))["elapsed"]
t_stri <- system.time(stri_length(x))["elapsed"]
cat(sprintf("base::nchar       %.3fs\nstringi::stri_length %.3fs\nfast.string::fnchar %.3fs\n", t_base, t_stri, t_fast))
record("fnchar", "base::nchar", t_base, t_fast)

t_base <- system.time(base::chartr("aeiou", "AEIOU", x))["elapsed"]
t_fast <- system.time(fast.string::fchartr("aeiou", "AEIOU", x))["elapsed"]
cat(sprintf("base::chartr      %.3fs\nfast.string::fchartr %.3fs\n", t_base, t_fast))
record("fchartr", "base::chartr", t_base, t_fast)
```

`fsubstr` is the one operation in this whole chapter where parallelism
doesn't help: base R's `substr()` is already close to memcpy speed for a
fixed-width extraction, so there's little headroom — consistent with the
~1x result in @tbl-summary. The win shows up on operations with real
per-element branching cost (`trimws`, `chartr`) or where base R's
implementation does something surprisingly expensive (`nchar`).

## Dates

```{r}
#| label: bench-dates
dates <- as.Date(sample(seq(as.Date("1950-01-01"), as.Date("2024-12-31"), by = "day"), n, replace = TRUE))
dates[na_idx] <- NA
date_strings_iso <- fast.string::format_date(dates, "iso")

t_base <- system.time(format(dates, "%Y-%m-%d"))["elapsed"]
t_fast <- system.time(fast.string::format_date(dates, "iso"))["elapsed"]
cat(sprintf("base format.Date  %.3fs\nfast.string::format_date %.3fs\n", t_base, t_fast))
record("format_date", "format.Date", t_base, t_fast)

t_base <- system.time(base::as.Date(date_strings_iso, format = "%Y-%m-%d"))["elapsed"]
t_fast <- system.time(fast.string::fas.Date(date_strings_iso, "iso"))["elapsed"]
cat(sprintf("base::as.Date     %.3fs\nfast.string::fas.Date %.3fs\n", t_base, t_fast))
record("fas.Date", "as.Date", t_base, t_fast)
```

## Phonetic blocking keys

`soundex()`/`nysiis()` have no base R equivalent; `RecordLinkage::soundex()`
is the closest comparison (algorithm details and why the two can disagree
on specific inputs are in @sec-theory). `RecordLinkage::soundex()` runs a
single timed pass rather than `microbenchmark`'s repeated runs — at 3M
rows it's the single slowest operation in this chapter, and repeating it
five times buys little additional statistical confidence for a lot of
extra render time.

```{r}
#| label: bench-phonetic
names_x <- stri_rand_strings(n, sample(3:15, n, replace = TRUE), pattern = "[A-Z]")
names_x[na_idx] <- NA_character_

t_fast_sx <- system.time(fast.string::soundex(names_x))["elapsed"]
t_fast_ny <- system.time(fast.string::nysiis(names_x))["elapsed"]
cat(sprintf("fast.string::soundex %.3fs\nfast.string::nysiis  %.3fs\n", t_fast_sx, t_fast_ny))

if (has_recordlinkage) {
    t_rl_sx <- system.time(RecordLinkage::soundex(names_x))["elapsed"]
    cat(sprintf("RecordLinkage::soundex %.3fs\n", t_rl_sx))
    record("soundex", "RecordLinkage::soundex", t_rl_sx, t_fast_sx)
} else {
    cat("RecordLinkage not installed — skipping cross-package soundex comparison.\n")
}
```

`nysiis()` is consistently slower than `soundex()` here — not a
parallelism gap (both run through the same dispatch), but because NYSIIS's
rule table is genuinely heavier per element (leading/trailing transforms
plus a stateful per-letter pass, vs. Soundex's single fixed lookup), as
described in @sec-theory.

`double_metaphone()` and `caverphone()` are considerably heavier rulesets
again (Double Metaphone's branching is the most involved of the four;
Caverphone 2.0 is a long fixed chain of substring transforms), so absolute
throughput is lower — `phonics::metaphone()`/`phonics::caverphone()` are
shown alongside as "another phonetic-coding R function over this vector"
throughput comparisons, not as algorithm-equivalent baselines (see the
note at the top of this chapter):

```{r}
#| label: bench-phonetic-v2
t_fast_dm <- system.time(fast.string::double_metaphone(names_x))["elapsed"]
t_fast_cp <- system.time(fast.string::caverphone(names_x))["elapsed"]
cat(sprintf("fast.string::double_metaphone %.3fs\nfast.string::caverphone       %.3fs\n",
            t_fast_dm, t_fast_cp))

if (has_phonics) {
    t_ph_mp <- system.time(phonics::metaphone(names_x))["elapsed"]
    t_ph_cp <- system.time(phonics::caverphone(names_x, maxCodeLen = 10))["elapsed"]
    cat(sprintf("phonics::metaphone   %.3fs\nphonics::caverphone  %.3fs\n", t_ph_mp, t_ph_cp))
    record("double_metaphone", "phonics::metaphone (different algorithm)", t_ph_mp, t_fast_dm)
    record("caverphone", "phonics::caverphone (different revision)", t_ph_cp, t_fast_cp)
} else {
    cat("phonics not installed — skipping cross-package comparison.\n")
}
```

## Fuzzy string similarity

```{r}
#| label: bench-jaro
t_fast <- system.time(fast.string::jaro_winkler(names_x, rev(names_x)))["elapsed"]
cat(sprintf("fast.string::jaro_winkler %.3fs (%s pairs, no base equivalent)\n",
            t_fast, format(n, big.mark = ",", scientific = FALSE)))

if (has_stringdist) {
    t_sd <- system.time(1 - stringdist::stringdist(names_x, rev(names_x), method = "jw", p = 0.1))["elapsed"]
    cat(sprintf("stringdist::stringdist(jw, p=0.1) %.3fs\n", t_sd))
    record("jaro_winkler", "stringdist::stringdist(jw)", t_sd, t_fast)
}
if (has_recordlinkage) {
    t_rl <- system.time(RecordLinkage::jarowinkler(names_x, rev(names_x)))["elapsed"]
    cat(sprintf("RecordLinkage::jarowinkler %.3fs\n", t_rl))
    record("jaro_winkler", "RecordLinkage::jarowinkler", t_rl, t_fast)
}
```

```{r}
#| label: bench-jaro-correctness
# All three should agree numerically (RecordLinkage's defaults and
# stringdist with p = 0.1 both implement the same standard Jaro-Winkler).
a3 <- c("MARTHA", "DWAYNE", "DIXON")
b3 <- c("MARHTA", "DUANE", "DICKSONX")
fast.string::jaro_winkler(a3, b3)
if (has_recordlinkage) RecordLinkage::jarowinkler(a3, b3)
if (has_stringdist) 1 - stringdist::stringdist(a3, b3, method = "jw", p = 0.1)
```

`jaro_winkler_matrix()` is $O(n \times m)$ by construction, so a 3M x 3M
matrix is neither a realistic workload nor something most machines could
hold in memory — a 3,000 x 3,000 block (9M cell comparisons) is a more
representative blocking-table size:

```{r}
#| label: bench-jaro-matrix
m <- 3000
a_small <- names_x[seq_len(m)]
b_small <- names_x[(n - m + 1):n]
t_fast <- system.time(fast.string::jaro_winkler_matrix(a_small, b_small))["elapsed"]
cat(sprintf("fast.string::jaro_winkler_matrix: %s x %s (%s cells) in %.3fs\n",
            m, m, format(m * m, big.mark = ","), t_fast))
```

`jaro_winkler_tokens()` does more work per pair (tokenising, an internal
similarity matrix per row, greedy assignment, plus the collapsed
comparison — @sec-theory) than plain `jaro_winkler()`, so it's
meaningfully slower per comparison. No package in this comparison set has
a token-reorder-aware equivalent, so this is absolute throughput only:

```{r}
#| label: bench-jaro-tokens
cap1 <- function(s) paste0(toupper(substr(s, 1, 1)), substr(s, 2, nchar(s)))
make_multitoken_names <- function(n) {
    paste(
        cap1(stri_rand_strings(n, sample(3:9, n, replace = TRUE), pattern = "[a-z]")),
        cap1(stri_rand_strings(n, sample(3:9, n, replace = TRUE), pattern = "[a-z]")),
        cap1(stri_rand_strings(n, sample(3:9, n, replace = TRUE), pattern = "[a-z]"))
    )
}
tok_a <- make_multitoken_names(n)
tok_b <- make_multitoken_names(n)
t_plain  <- system.time(fast.string::jaro_winkler(tok_a, tok_b))["elapsed"]
t_tokens <- system.time(fast.string::jaro_winkler_tokens(tok_a, tok_b))["elapsed"]
cat(sprintf("fast.string::jaro_winkler        %.3fs (%s pairs)\n", t_plain, format(n, big.mark = ",", scientific = FALSE)))
cat(sprintf("fast.string::jaro_winkler_tokens %.3fs (%.1fx the plain-comparison cost)\n", t_tokens, t_tokens / t_plain))
```

## Edit distance

```{r}
#| label: bench-editdist
t_fast_lv <- system.time(fast.string::levenshtein(names_x, rev(names_x)))["elapsed"]
t_fast_osa <- system.time(fast.string::osa_distance(names_x, rev(names_x)))["elapsed"]
t_fast_dl <- system.time(fast.string::damerau_levenshtein(names_x, rev(names_x)))["elapsed"]
cat(sprintf("fast.string::levenshtein         %.3fs\nfast.string::osa_distance        %.3fs\nfast.string::damerau_levenshtein %.3fs\n",
            t_fast_lv, t_fast_osa, t_fast_dl))

if (has_stringdist) {
    t_sd_lv <- system.time(stringdist::stringdist(names_x, rev(names_x), method = "lv"))["elapsed"]
    t_sd_osa <- system.time(stringdist::stringdist(names_x, rev(names_x), method = "osa"))["elapsed"]
    t_sd_dl <- system.time(stringdist::stringdist(names_x, rev(names_x), method = "dl"))["elapsed"]
    cat(sprintf("stringdist (lv) %.3fs\nstringdist (osa) %.3fs\nstringdist (dl) %.3fs\n", t_sd_lv, t_sd_osa, t_sd_dl))
    record("levenshtein", "stringdist::stringdist(lv)", t_sd_lv, t_fast_lv)
    record("osa_distance", "stringdist::stringdist(osa)", t_sd_osa, t_fast_osa)
    record("damerau_levenshtein", "stringdist::stringdist(dl)", t_sd_dl, t_fast_dl)
}

# hamming() needs equal-length pairs to avoid the all-Inf degenerate case
names_eqlen <- fast.string::fsubstr(names_x, 1, 8)
t_fast_hm <- system.time(fast.string::hamming(names_eqlen, rev(names_eqlen)))["elapsed"]
cat(sprintf("fast.string::hamming %.3fs\n", t_fast_hm))
if (has_stringdist) {
    t_sd_hm <- system.time(stringdist::stringdist(names_eqlen, rev(names_eqlen), method = "hamming"))["elapsed"]
    cat(sprintf("stringdist (hamming) %.3fs\n", t_sd_hm))
    record("hamming", "stringdist::stringdist(hamming)", t_sd_hm, t_fast_hm)
}
```

`fuzzy_match()` computes the same best-match result without retaining the
full score matrix. The output matrix below is modest for documentation, but
its allocation grows as `length(query) * length(table)` while lookup output
remains one integer per query.

```{r}
#| label: bench-fuzzy-lookup
lookup_queries <- names_x[seq_len(100L)]
lookup_table <- names_x[1001:2000]
t_lookup <- system.time(
    fuzzy_match(lookup_queries, lookup_table, use_bytes = TRUE)
)["elapsed"]
t_matrix <- system.time(
    max.col(jaro_winkler_matrix(lookup_queries, lookup_table),
            ties.method = "first")
)["elapsed"]
cat(sprintf("fuzzy_match %.3fs; materialized matrix %.3fs\n",
            t_lookup, t_matrix))
```

```{r}
#| label: bench-editdist-correctness
# Each edit-distance implementation should agree with its stringdist method.
a3 <- c("kitten", "Saturday", "flaw")
b3 <- c("sitting", "Sunday", "lawn")
fast.string::levenshtein(a3, b3)
if (has_stringdist) stringdist::stringdist(a3, b3, method = "lv")
fast.string::osa_distance("ca", "abc")
if (has_stringdist) stringdist::stringdist("ca", "abc", method = "osa")
fast.string::damerau_levenshtein("ca", "abc")
if (has_stringdist) stringdist::stringdist("ca", "abc", method = "dl")
```

## Q-gram set similarity

```{r}
#| label: bench-qgram
t_fast_jac <- system.time(fast.string::jaccard_index(names_x, rev(names_x)))["elapsed"]
cat(sprintf("fast.string::jaccard_index %.3fs\n", t_fast_jac))
if (has_stringdist) {
    t_sd_jac <- system.time(stringdist::stringsim(names_x, rev(names_x), method = "jaccard"))["elapsed"]
    cat(sprintf("stringdist::stringsim(jaccard) %.3fs\n", t_sd_jac))
    record("jaccard_index", "stringdist::stringsim(jaccard)", t_sd_jac, t_fast_jac)
}

t_fast_dice <- system.time(fast.string::dice_coefficient(names_x, rev(names_x)))["elapsed"]
t_fast_tver <- system.time(fast.string::tversky_index(names_x, rev(names_x), alpha = 0.3, beta = 0.7))["elapsed"]
cat(sprintf("fast.string::dice_coefficient %.3fs (no direct stringdist equivalent)\n", t_fast_dice))
cat(sprintf("fast.string::tversky_index    %.3fs (no direct stringdist equivalent)\n", t_fast_tver))
```

## fuzzywuzzy-style ratios

No R package implements fuzzywuzzy's specific scoring (Ratcliff/Obershelp
matching blocks); this is absolute throughput only.

```{r}
#| label: bench-fuzzywuzzy
t_ratio   <- system.time(fast.string::fuzz_ratio(tok_a, tok_b))["elapsed"]
t_partial <- system.time(fast.string::fuzz_partial_ratio(tok_a, tok_b))["elapsed"]
t_tsort   <- system.time(fast.string::fuzz_token_sort_ratio(tok_a, tok_b))["elapsed"]
t_tset    <- system.time(fast.string::fuzz_token_set_ratio(tok_a, tok_b))["elapsed"]
cat(sprintf(
    "fast.string::fuzz_ratio            %.3fs\nfast.string::fuzz_partial_ratio    %.3fs\nfast.string::fuzz_token_sort_ratio %.3fs\nfast.string::fuzz_token_set_ratio  %.3fs\n",
    t_ratio, t_partial, t_tsort, t_tset
))
```

## The small-n crossover

@sec-theory claims every parallel function applies a size threshold and
falls back to single-threaded execution below it, *because* thread-pool
dispatch overhead would otherwise exceed the time saved. Here's that claim
checked directly, at `n = 20` — small enough that the claim predicts
`fast.string` will be *no faster, possibly slower* than base R:

```{r}
#| label: bench-small-n
small <- stri_rand_strings(20, sample(5:15, 20, replace = TRUE))
print(microbenchmark(
    base = base::grepl("[0-9]", small),
    fast = fast.string::fgrepl("[0-9]", small),
    times = 50
))
```

At this scale the two are within noise of each other (sometimes base R
even wins outright) — exactly what the size-threshold design predicts, and
the reason the rest of this chapter uses 3M rows rather than 20: the
benefit this package offers only shows up once there's enough work to
amortise parallel dispatch.

## Summary

```{r}
#| label: tbl-summary
#| tbl-cap: !expr sprintf("Observed speedups, n = %s, %d cores, 1 run per cell unless noted (this machine — yours will vary).", format(n, big.mark = ",", scientific = FALSE), parallel::detectCores())
summary_df <- do.call(rbind, results)
summary_df$baseline_s <- round(summary_df$baseline_s, 3)
summary_df$fast_s     <- round(summary_df$fast_s, 3)
summary_df$speedup    <- paste0(round(summary_df$speedup, 1), "x")
knitr::kable(
    summary_df,
    row.names = FALSE,
    col.names = c("Operation", "Baseline", "Baseline (s)", "fast.string (s)", "Speedup")
)
```

A few patterns worth drawing out explicitly:

- The biggest wins are on operations with **real per-element work** —
  regex matching/substitution, trimming, phonetic encoding — where
  parallelising actual computation pays off directly.
- The smallest (or negative) wins are on operations base R already does
  close to optimally for a single core (`substr()`); parallelism helps
  least when there's least work to spread around.
- Cross-package comparisons (`stringdist`, `RecordLinkage`) confirm
  *correctness* (the Jaro-Winkler numbers agree to floating-point
  precision) as much as they confirm speed — agreement there is the more
  important result, since a faster-but-wrong implementation isn't a
  useful trade.
