# Troubleshooting & FAQ {#sec-troubleshooting}

```{r}
#| label: setup
#| include: false
source("_setup.R")
```

## Installation

### "could not find tools necessary to compile a package" / install fails

`fast.string` compiles C++ source on install (PCRE2 bindings, RE2,
RcppParallel). You need a working toolchain:

- **Windows**: install [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
  matching your R version, and make sure it's on `PATH` (the Rtools
  installer offers to do this). Verify with
  `pkgbuild::has_build_tools(debug = TRUE)`.
- **macOS**: install Xcode command-line tools (`xcode-select --install`).
- **Linux**: a C++14-capable compiler (gcc/clang) plus the PCRE2
  development headers — `libpcre2-dev` on Debian/Ubuntu,
  `pcre2-devel` on Fedora/RHEL.

If `pkgbuild::has_build_tools(debug = TRUE)` returns `FALSE`, the
underlying compiler invocation it prints is the fastest way to see exactly
what's missing.

### Reinstalling after pulling new source changes

If you're developing against a local checkout rather than installing a
release, `devtools::load_all()` recompiles changed `src/*.cpp` files
automatically; a plain `library(fast.string)` won't pick up source changes
without reinstalling first.

## NA, empty strings, and encoding

### Why does my function return `NA` for a non-`NA` input?

Every function in this package returns `NA` for `NA` input — but a few
functions also return `NA` for specific *valid-looking but unparseable*
inputs:

- `soundex()`/`nysiis()`/`double_metaphone()` return `NA` for strings with
  **no alphabetic characters at all** (e.g. `"123"`, `""`) — there's no
  phonetic code for a string with nothing to encode phonetically.
  `caverphone()` is the exception: it always returns a 10-character code
  for non-`NA` input (even `""`, which codes as `"1111111111"`), matching
  its reference implementation's behaviour.
- `fas.Date()` returns `NA` for any string that doesn't match the
  specified `format` *exactly* (wrong length, wrong separator position,
  non-digit where a digit is expected) or has an out-of-range month/day.
  Unlike `base::as.Date()`, it does **not** attempt multiple formats or
  partial parsing.

```{r}
soundex(c("Smith", "123", "", NA))
fas.Date(c("2024-06-18", "2024/06/18", "2024-13-01"), "iso")  # 2nd: wrong separator, 3rd: month 13
```

### `fchartr()` silently fell back to `base::chartr()` — why?

`fchartr()`'s fast path is a flat 256-entry byte lookup table, which only
works correctly when every character in `old` and `new` is single-byte
(ASCII). If either argument contains a multi-byte UTF-8 character (an
accented letter, a curly quote, an emoji), the function detects that via
a byte-count vs. char-count comparison and transparently delegates to
`base::chartr()` instead — same result, just without the speedup, and
without you needing to know in advance.

```{r}
fchartr("e", "é", "test")  # "é" is multi-byte — falls back to base::chartr(), still correct
```

### `fnchar()` bytes vs. chars

`type = "bytes"` counts bytes; `type = "chars"` counts Unicode codepoints.
These differ for any non-ASCII input — get this backwards and a
fixed-width field extraction (`fsubstr()`) on multi-byte data will be off
by however many multi-byte characters are in the string.

```{r}
fnchar("café", type = "bytes")  # 5 — "é" is 2 bytes in UTF-8
fnchar("café", type = "chars") # 4 — 4 visible characters
```

## Threading

### I set `nthreads = 1` for one call, and now *every* call is slow

This is real, and worth knowing about explicitly:
`RcppParallel::setThreadOptions()` configures the underlying TBB thread
pool **process-wide**, not just for the call that set it. Passing
`nthreads` to any function that accepts it changes the thread count for
*every subsequent parallel call in the session*, until something resets
it.

```{r}
#| label: thread-persistence
x <- replicate(2e6, paste(sample(letters, 10, TRUE), collapse = ""))

t1 <- system.time(jaro_winkler(x, rev(x)))["elapsed"]
jaro_winkler("a", "a", nthreads = 1)  # a tiny, throwaway call — but it sets the pool to 1 thread
t2 <- system.time(jaro_winkler(x, rev(x)))["elapsed"]
RcppParallel::setThreadOptions(numThreads = "auto")
t3 <- system.time(jaro_winkler(x, rev(x)))["elapsed"]

cat(sprintf("before:            %.2fs\nafter nthreads=1:  %.2fs\nafter reset:       %.2fs\n", t1, t2, t3))
```

That single-row, throwaway `nthreads = 1` call made every later
full-vector call run single-threaded too — an 8x+ slowdown on this
16-core machine, silently, with no warning. If you need a one-off thread
count for a specific call and want every other call in the session
unaffected, reset explicitly afterward:

```r
RcppParallel::setThreadOptions(numThreads = "auto")
```

A handful of functions (`ftrimws()`, `fsubstr()`, `fnchar()`, `fchartr()`,
the `format_date()` family, `soundex()`, `nysiis()`, `double_metaphone()`,
`caverphone()`) don't take `nthreads` at all — they always run at whatever
the process-wide setting currently is, so they're affected by this too
even though they have no argument that looks like it would be.

### Does `nthreads` help on a machine with few cores?

Not much below 4 cores, and the package's own size thresholds
(@sec-theory) mean it won't even try to parallelise small inputs
regardless. If you're running inside a container with a CPU limit lower
than `parallel::detectCores()` reports (common in CI and some cloud
environments), set `nthreads` explicitly to the actual limit — TBB doesn't
know about cgroup CPU quotas and will otherwise oversubscribe.

## Regex behaviour

### My pattern works in `grepl(perl = TRUE)` but `fgrepl()` gave a `cli::cli_inform()` message and fell back

That's the intended behaviour, not a bug: the pattern uses PCRE-only
syntax (lookaround, atomic groups, possessive quantifiers, named
backreferences — the exact list is in @sec-theory), and `fast.string`
delegates to `base::grepl(perl = TRUE)` automatically rather than risk a
subtly different result from its own PCRE2 binding on syntax it hasn't
specifically validated. The result is correct either way; you're just not
getting the parallel speedup for that specific call. Pass `perl = TRUE`
explicitly to skip the syntax check (only do this if you've confirmed the
pattern behaves identically under `fast.string`'s PCRE2 path for your
actual data).

### Catastrophic backtracking — am I safe?

Only partially. RE2 (used for `fixed = TRUE`) is immune by construction.
PCRE2 (used for general regex) is **not** — it's still a backtracking
engine, same risk class as base R's `perl = TRUE`. If you're matching
untrusted patterns against untrusted input, that risk exists regardless of
which package you use; this package doesn't add to it, but doesn't remove
it either. Prefer `fixed = TRUE` whenever the pattern is actually a literal
string — beyond being immune to backtracking blowup, it's also the fastest
path in the package.

## Getting help

- Found a bug, or a discrepancy between `fast.string` and base R's
  behaviour on some input? Open an issue:
  <https://github.com/KyleHaynes/anthropic_fast/issues>.
- For the underlying algorithms, @sec-theory has the full detail behind
  every function's behaviour.
- For exact argument-by-argument documentation, @sec-reference.
